### 2024-11-12 paper/paper.qmd
Updated scripts

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(knitr)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada. The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [@citeProjectHammerData]. This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.
Project Hammer aims to support regulatory efforts by providing data on price transparency, making it suitable for academic research and potential legal analysis. Following data processing and cleaning procedures outlined in Section 3.2, the dataset was structured to allow an analysis of historical price changes and vendor-specific price patterns.
The dataset contains 1,996,969 rows and 5 columns, with variables capturing product (product name), vendor_name (name of the grocery retailer), price_current (current product price at the time of recording), price_old (previous recorded price for the product), and price_difference (difference between the current and old prices). Price data is recorded in Canadian dollars and captures a broad range of grocery items from various categories, including fresh produce, dairy, pantry staples, and household items.
To ensure the quality and consistency of the data, we focused on removing missing values and imputing outlier prices when necessary. For this analysis, we excluded entries with extreme price fluctuations beyond three standard deviations, as these could represent temporary discounts or data recording anomalies. This cleaning process provides a balanced and accurate reflection of grocery pricing, allowing for robust statistical analysis on vendor-specific trends and average price comparisons across product types.
By narrowing the scope to recent data, Project Hammer's dataset captures real-time fluctuations in grocery pricing, providing timely insights into potential market dynamics and consumer price impacts. The analysis allows for future investigation of temporal price trends, vendor-based price differences, and, more broadly, consumer affordability in a market facing rising cost-of-living pressures.

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

Our modeling approach seeks to explore and quantify the relationship between previous grocery prices, vendor identities, and price differences in the current prices observed within Canadian grocery stores. This analysis employs a Bayesian linear regression model implemented via the `stan
_glm` function in the `rstanarm` package to examine how factors such as historical prices, vendor differences, and observed price changes impact current prices for various grocery items.

In this model, `price_current` serves as the response variable, while `price_old`, `vendor_name`, and `price_difference` act as predictor variables. The linear regression model assumes a Gaussian distribution for the response variable `price_current`, allowing for a straightforward interpretation of the estimated parameters.

## Model set-up

The model includes the following predictor variables:

- Previous Price (`price_old`): The price of the product in a previous time period.

- Vendor (`vendor_name`): A categorical variable representing the grocery store chain selling the product, capturing vendor-specific pricing differences.

- Price Difference (`price_difference`): The difference between the current and previous price, which may indicate market or vendor-specific pricing adjustments.

The model can be represented mathematically as follows:

\begin{align*}
    y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
    \mu_i &= \beta_0 + \beta_1 \cdot \text{Previous Price}_i + \beta_2 \cdot \text{Vendor}_i + \beta_3 \cdot \text{Price Difference}_i \\
    \epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

\textbf{Where:}
\begin{itemize}
    \item $\beta_0$ is the intercept term, representing the baseline estimage of `price_current`
    \item $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients representing the effects of `price_old`, `vendor_name` and `price_difference` on `price_current`.
    \item $\sigma^2$represents the variance of the error term, capturing unexplained variability in current prices.
\end{itemize}

The model is executed in \texttt{R} [@citeR] using the \texttt{rstanarm} package [@rstanarm],  with priors set to regularize the estimates and prevent overfitting. Specifically, we use a normal prior for the coefficients, centered at zero with moderate variance, to ensure stable estimates without overly restrictive assumptions.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

The results of our analysis of the Project Hammer dataset explain important findings about pricing trends across major Canadian grocery vendors. By examining factors such as prior pricing (`price_old`), vendor identities (`vendor_name`), and recent price changes (`price_difference`), we assess how these variables affect current prices (`price_current`). The following sections summarize key findings, supported by visualizations that highlight trends and relationships within the data.

## Vendor-Specific Price Trends

We begin by examining average current prices (`price_current`) across different grocery vendors to identify potential pricing variations in the Canadian grocery market.

```{r}
#### Read data ####
# Load the cleaned dataset
analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv", show_col_types = FALSE)

# Generate plot for Average Current Price by Vendor
vendor_price_plot <- analysis_data %>%
  group_by(vendor_name) %>%
  summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Average Current Price by Vendor",
    x = "Vendor",
    y = "Average Current Price (CAD)"
  ) +
  theme_minimal()

# Display the plot
vendor_price_plot

```

Figure 1: Average Current Price by Vendor. This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD). Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.

## Price Difference Analysis

We further analyze `price_difference`, which captures the change between the current price (`price_current`) and the previous price (`price_old`). This variable provides insights into recent price adjustments across vendors.

```{r}
# Average price difference by vendor
price_diff_plot <- analysis_data %>%
  group_by(vendor_name) %>%
  summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(
    title = "Average Price Difference by Vendor",
    x = "Vendor",
    y = "Average Price Difference (CAD)"
  ) +
  theme_minimal()
price_diff_plot

```

Figure 2: Average Price Difference by Vendor. This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments. Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.

## Relationship Between Previous and Current Prices

We assess the relationship between previous prices (`price_old`) and current prices (`price_current`) to understand price consistency and potential price elasticity across products.

```{r}
# Scatter plot of previous price vs. current price
price_relationship_plot <- ggplot(analysis_data, aes(x = price_old, y = price_current)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Relationship Between Previous and Current Prices",
    x = "Previous Price (CAD)",
    y = "Current Price (CAD)"
  ) +
  theme_minimal()
price_relationship_plot

```
Figure 3: Relationship Between Previous and Current Prices. This scatter plot with a linear regression line shows the correlation between previous and current prices. A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.

## Distribution of Price Differences

To further explore pricing dynamics, we examine the distribution of `price_difference`, capturing both price increases and decreases across the dataset.

```{r}
# Calculate mean and standard deviation of price_difference
mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)

# Filter out extreme values beyond 3 standard deviations
filtered_data <- analysis_data %>%
  filter(price_difference >= (mean_diff - 3 * sd_diff),
         price_difference <= (mean_diff + 3 * sd_diff))

# Plot histogram with filtered data
price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(
    title = "Distribution of Price Differences (Filtered)",
    x = "Price Difference (CAD)",
    y = "Frequency"
  ) +
  theme_minimal()
price_difference_hist

```

Figure 4: Distribution of Price Differences (Filtered). This histogram illustrates the frequency of price changes across products, with extreme values beyond three standard deviations removed for clarity. The concentration around zero suggests that small price adjustments are most common, while the presence of both negative and positive values indicates instances of both price increases and decreases. This distribution provides insight into the stability and variability of pricing across vendors.
## Model Evaluation

To evaluate our model’s performance, we assess the predictive accuracy of the Bayesian linear regression model on the test data. The model’s Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) provide insights into the model’s ability to predict `price_current` based on `price_old`, `vendor_name`, and `price_difference`.

```{r}
# Calculate MAE and RMSE for model evaluation
mae <- mean(abs(predictions - analysis_data_test$price_current))
rmse <- sqrt(mean((predictions - analysis_data_test$price_current)^2))

evaluation_metrics <- data.frame(
  Metric = c("Mean Absolute Error (MAE)", "Root Mean Squared Error (RMSE)"),
  Value = c(round(mae, 2), round(rmse, 2))
)

# Display the results table
kable(evaluation_metrics, col.names = c("Evaluation Metric", "Value"))

print(evaluation_metrics)

```

Table 1: Model Evaluation Metrics. The table displays the MAE and RMSE for the model, reflecting the average and squared prediction errors, respectively. Lower values indicate better predictive accuracy, suggesting that the model effectively captures the relationship between previous prices, vendor identities, and price differences.

These results demonstrate that vendor identity and previous prices are significant predictors of current prices in the Canadian grocery sector, with implications for understanding competitive strategies and consumer price sensitivity in the market.

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-11-12 scripts/00-simulate_data.R

@@ -1,52 +1,42 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Simulates a dataset for Project Hammer, which aims to monitor grocery prices 
+#          across vendors and products to analyze market competition in Canada.
+# Author: Elizabeth Luong and Abdullah Motasim
+# Date: 14 November 2024
+# Contact: elizabethh.luong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Pre-requisites: The `tidyverse` package must be installed.
+# Note: Run this script in an R project workspace if needed.
 
 #### Workspace setup ####
 library(tidyverse)
 set.seed(853)
 
-
 #### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
+
+# Vendor names (8 vendors as described)
+vendors <- c("Voila", "T&T", "Loblaws", "No Frills", "Metro", "Galleria", "Walmart", "Save-On-Foods")
+
+# Product categories (for variety in products)
+product_categories <- c("Dairy", "Produce", "Bakery", "Meat", "Beverages", "Snacks", "Frozen Foods", "Household Items")
+
+# Product names (a simple list of items in each category)
+product_names <- c("Milk", "Bread", "Apple", "Chicken Breast", "Soda", "Chips", "Frozen Pizza", "Detergent")
+
+# Generate a date range for observation dates
+date_range <- seq(as.Date("2024-02-28"), as.Date("2024-11-01"), by = "day")
+
+# Create a simulated dataset by randomly assigning values to each variable
+hammer_data <- tibble(
+  product_id = 1:1000, # Unique ID for each product
+  product_name = sample(product_names, size = 1000, replace = TRUE),
+  category = sample(product_categories, size = 1000, replace = TRUE),
+  vendor = sample(vendors, size = 1000, replace = TRUE),
[…]

### 2024-11-12 scripts/01-test_simulated_data.R

@@ -1,89 +1,141 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Tests the structure and validity of the simulated Project Hammer dataset, 
+#          which includes grocery price data across vendors and products.
+# Author: Elizabeth Luong and Abdullah Motasim
+# Date: 14 November 2024
+# Contact: elizabethh.luong@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+#   - The `tidyverse` package must be installed and loaded
+#   - 00-simulate_data.R must have been run
+#   - Make sure you are in the correct project directory.
 
 
 #### Workspace setup ####
 library(tidyverse)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+# Load the dataset
+hammer_data <- read_csv("data/00-simulated_data/project_hammer_simulated_data.csv")
 
 # Test if the data was successfully loaded
-if (exists("analysis_data")) {
+if (exists("hammer_data")) {
   message("Test Passed: The dataset was successfully loaded.")
 } else {
   stop("Test Failed: The dataset could not be loaded.")
 }
 
+#### Test data structure ####
 
-#### Test data ####
-
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
+# Check if the dataset has 1000 rows (as per the simulated data)
+if (nrow(hammer_data) == 1000) {
+  message("Test Passed: The dataset has 1000 rows.")
 } else {
-  stop("Test Failed: The dataset does not have 151 rows.")
+  stop("Test Failed: The dataset does not have 1000 rows.")
 }
 
-# Check if the dataset has 3 columns
-if (ncol(analysis_data) == 3) {
-  message("Test Passed: The dataset has 3 columns.")
+# Check if the dataset has 7 columns
+if (ncol(hammer_data) == 7) {
+  message("Test Passed: The dataset has 7 columns.")
 } else {
-  stop("Test Failed: The dataset does not have 3 columns.")
[…]

### 2024-11-12 scripts/02-download_data.R

@@ -1,26 +1,75 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Loads data from uploaded files and tests the structure and validity 
+#          of the Project Hammer dataset, specifically checking price consistency.
+# Author: Elizabeth Luong and Abdullah Motasim
+# Date: 14 November 2024
+# Contact: elizabethh.luong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: The `tidyverse` package must be installed.
 
 
 #### Workspace setup ####
-library(opendatatoronto)
 library(tidyverse)
-# [...UPDATE THIS...]
 
-#### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
+# Define the directory to store files and ensure it exists
+file_directory <- "C:/Users/eliza/Downloads/hammer-3-compressed"
+if (!dir.exists(file_directory)) {
+  dir.create(file_directory, recursive = TRUE)
+}
 
+# Define file names and their URLs (replace <URL> with the actual URL for each file)
+files <- list(
+  "average_price_by_vendor.csv" = "<URL_to_average_price_by_vendor.csv>",
+  "price_trends_over_time.csv" = "<URL_to_price_trends_over_time.csv>",
+  "largest_price_fluctuations.csv" = "<URL_to_largest_price_fluctuations.csv>",
+  "price_comparison_specific_product.csv" = "<URL_to_price_comparison_specific_product.csv>",
+  "historical_price_specific_product.csv" = "<URL_to_historical_price_specific_product.csv>",
+  "price_changes.csv" = "<URL_to_price_changes.csv>",
+  "unique_products_per_vendor.csv" = "<URL_to_unique_products_per_vendor.csv>",
+  "product.csv" = "<URL_to_product.csv>"
+)
 
+# Download files if they do not already exist
+for (file_name in names(files)) {
+  file_path <- file.path(file_directory, file_name)
+  if (!file.exists(file_path)) {
+    download.file(url = files[[file_name]], destfile = file_path, mode = "wb")
+    message(paste("Downloaded:", file_name))
[…]

### 2024-11-12 scripts/03-clean_data.R

@@ -1,44 +1,47 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the Project Hammer data recorded by various sources and formats it for analysis
+# Author: Elizabeth Luong and Abdullah Motasim
+# Date: 14 November 2024
+# Contact: elizabethh.luong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: The `tidyverse` and `janitor` packages must be installed
+# Any other information needed? Ensure the data files have been downloaded and are in the specified directory
 
 #### Workspace setup ####
 library(tidyverse)
+library(janitor)
+
+# Define the directory where data files are located
+file_directory <- "C:/Users/eliza/Downloads/hammer-3-compressed"
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
 
+# Load raw data from the price_changes file
+raw_data <- read_csv(file.path(file_directory, "price_changes.csv"))
+
+# Begin cleaning and formatting the data
 cleaned_data <-
   raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
+  janitor::clean_names() |>  # Clean column names to snake_case
+  select(product_name, vendor, current_price, old_price, price_change) |>  # Select relevant columns
   mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
+    # Clean current_price and old_price by removing non-numeric characters
+    current_price = as.numeric(gsub("[^0-9.]", "", current_price)),
+    old_price = as.numeric(gsub("[^0-9.]", "", old_price))
   ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
[…]

### 2024-11-12 scripts/04-test_analysis_data.R

@@ -1,69 +1,76 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests the structure and validity of the cleaned Project Hammer dataset
+# Author: Elizabeth Luong and Abdullah Motasim
+# Date: 14 November 2024
+# Contact: elizabethh.luong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: `tidyverse` and `testthat` packages must be installed
+# Any other information needed? Ensure `cleaned_price_changes.csv` is present in the specified directory
 
 #### Workspace setup ####
 library(tidyverse)
 library(testthat)
 
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
+# Load the cleaned dataset
+cleaned_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
 
+# Ensure `price_difference` is calculated correctly as `price_current - price_old`
+cleaned_data <- cleaned_data %>%
+  mutate(price_difference = price_current - price_old)
+
+# Identify rows with out-of-range prices
+out_of_range <- cleaned_data %>%
+  filter(price_current < 1 | price_current > 20 | price_old < 1 | price_old > 20)
+
+# Display the out-of-range entries for inspection
+print("Out-of-range values in 'price_current' or 'price_old':")
+print(out_of_range)
+
+# Optionally, filter data to include only rows within the $1 to $20 range
+cleaned_data <- cleaned_data %>%
+  filter(price_current >= 1, price_current <= 20, price_old >= 1, price_old <= 20)
 
 #### Test data ####
-# Test that the dataset has 151 rows - there are 151 divisions in Australia
-test_that("dataset has 151 rows", {
-  expect_equal(nrow(analysis_data), 151)
+
+# Test that the dataset has the expected number of columns (5 in this case)
+test_that("dataset has 5 columns", {
+  expect_equal(ncol(cleaned_data), 5)
 })
 
-# Test that the dataset has 3 columns
-test_that("dataset has 3 columns", {
-  expect_equal(ncol(analysis_data), 3)
[…]

### 2024-11-12 scripts/05-exploratory_data_analysis.R

@@ -1,37 +1,101 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Exploratory data analysis on cleaned Project Hammer data
+# Author: Elizabeth Luong and Abdullah Motasim
+# Date: 14 November 2024
+# Contact: elizabethh.luong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+#   - The `tidyverse` package must be installed and loaded
+#   - The cleaned dataset `cleaned_price_changes.csv` must be available
 
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+# Load the cleaned dataset
+analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+# General overview of the dataset
+glimpse(analysis_data)
+summary(analysis_data)
 
 
-#### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
-)
+#### Summary Statistics ####
+# Summary of numeric columns
+analysis_data %>%
+  select_if(is.numeric) %>%
+  summary()
+
+# Count unique values in categorical columns
+analysis_data %>%
+  select_if(is.character) %>%
+  summarise_all(~ n_distinct(.))
 
 
+#### Missing Values Analysis ####
+# Check for missing values in each column
+analysis_data %>%
+  summarise_all(~ sum(is.na(.)))
+
+
+#### Univariate Analysis ####
+# Distribution of current price
+ggplot(analysis_data, aes(x = price_current)) +
[…]

### 2024-11-12 scripts/06-model_data.R

@@ -1,37 +1,74 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Models Project Hammer data to predict current grocery prices using stan_glm with efficient memory handling.
+# Author: Elizabeth Luong and Abdullah Motasim
+# Date: 14 November 2024
+# Contact: elizabethh.luong@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: 
+#   - The `tidyverse`, `rstanarm`, `arrow`, `data.table` packages must be installed and loaded
+#   - The cleaned dataset must be available as `cleaned_price_changes.csv`
 
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+library(arrow)
+library(data.table)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+# Load the cleaned data with data.table for efficiency
+analysis_data <- fread("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
+
+# Split data into training and testing sets
+set.seed(853)
+train_indices <- sample(seq_len(nrow(analysis_data)), size = 0.8 * nrow(analysis_data))
+analysis_data_train <- analysis_data[train_indices]
+analysis_data_test <- analysis_data[-train_indices]
+
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
-
+# Define a Bayesian regression model to predict `price_current`
+model <- stan_glm(
+  formula = price_current ~ price_old + as.factor(vendor_name) + price_difference,
+  data = analysis_data_train,
+  family = gaussian(),
+  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
[…]

### 2024-11-13 paper/paper.qmd
Added visuals

@@
 library(knitr)
+library(rstanarm)
+library(arrow)
+library(data.table)
 ```
@@
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+The selection of a linear regression model for this analysis is based on the continuous nature of `price_current`, which allows us to quantify how previous pricing, vendor, and price changes influence current pricing.
+Existing economic theories suggest that previous prices and historical pricing patterns are often predictive of current pricing, especially in competitive retail markets where price sensitivity and vendor-specific strategies play significant roles.
+The inclusion of `price_old` as a predictor aligns with time-series economic models where past data points inform future values.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+Additionally, vendor differences, represented by `vendor_name`, capture potential competitive dynamics in the Canadian grocery market.
+Vendors may adopt distinct pricing strategies or respond differently to market conditions, and the inclusion of this categorical variable enables an analysis of these patterns across various major grocery chains.
+The `price_difference` variable, which represents recent price changes, provides an understanding of market adjustments and reflects fluctuations potentially influenced by external factors, such as supply chain constraints or inflation.
+
+The Bayesian approach was selected for its flexibility in incorporating prior information and its robustness in handling uncertainty within small or moderate sample sizes.
+Furthermore, Bayesian methods allow us to capture the posterior distributions of the parameters, which can be used to evaluate the strength and credibility of each predictor's effect on current prices.
+This approach also aligns with the objectives of Project Hammer by facilitating a transparent and probabilistically rigorous analysis of grocery pricing trends.
+
+In summary, this model helps explain the primary factors of current grocery prices in Canadian stores, accounting for both historical prices and vendor effects.
+By utilizing Bayesian framework, this analysis provides a clear quantification of the influence of each predictor while allowing for robust estimation and interpretability in the context of economic and competitive pricing strategies.
 
@@
 ```{r}
+
+#### Read data ####
+# Load the cleaned data with data.table for efficiency
+analysis_data <- fread("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
+
+# Split data into training and testing sets
+set.seed(853)
+train_indices <- sample(seq_len(nrow(analysis_data)), size = 0.8 * nrow(analysis_data))
+analysis_data_train <- analysis_data[train_indices]
+analysis_data_test <- analysis_data[-train_indices]
+### Model data ####
+# Define a Bayesian regression model to predict `price_current`
+
+### Model data ####
+# Define a Bayesian regression model to predict `price_current`
+model <- stan_glm(
+  formula = price_current ~ price_old + as.factor(vendor_name) + price_difference,
+  data = analysis_data_train,
+  family = gaussian(),
+  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_aux = exponential(rate = 1, autoscale = TRUE),
+  seed = 853
+)
+
+# Define the chunk size (adjust based on available memory)
+chunk_size <- 50000  
+num_chunks <- ceiling(nrow(analysis_data_test) / chunk_size)
+
+# Initialize a vector to store predictions
+predictions <- vector("numeric", nrow(analysis_data_test))
+
+# Loop through each chunk and generate predictions
+for (i in 1:num_chunks) {
+  start <- ((i - 1) * chunk_size) + 1
+  end <- min(i * chunk_size, nrow(analysis_data_test))
+
+  # Generate predictions for the current chunk
+  predictions[start:end] <- predict(model, newdata = analysis_data_test[start:end, ])
+}
+
 # Calculate MAE and RMSE for model evaluation
@@
 
-## First discussion point {#sec-first-point}
+## Interpretation of Grocery Pricing Dynamics
+This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset.
+Our Bayesian regression model reveals that prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`) play significant roles in determining current prices.
+The influence of previous prices indicates price continuity across time, suggesting that grocery items tend to exhibit relatively stable pricing patterns with minor adjustments.
+Vendor-specific effects highlight the role of retailer pricing strategies, with some vendors maintaining consistently higher or lower prices.
+These vendor variations may reflect differences in market positioning, operational costs, or competitive strategies, impacting price consistency across the grocery sector.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Additionally, the presence of both positive and negative values in the `price_difference` variable illustrates the prevalence of both price increases and decreases, underscoring the dynamic nature of grocery pricing in response to external market forces.
+External factors, such as supply chain disruptions, inflation, and seasonal demand fluctuations, likely contribute to the observed variability.
+This model highlights the complexity of pricing dynamics and the importance of historical pricing and vendor-specific factors in shaping current grocery prices across major Canadian retailers.
 
-## Second discussion point
+## Implications for Competition and Consumer Costs
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+The findings from Project Hammer have broader implications for both competition in the grocery sector and consumer expenses.
+Vendor-specific price differences suggest that some retailers adopt competitive pricing strategies to attract cost-sensitive consumers, while others may leverage brand loyalty or perceived quality to justify higher prices.
+In this context, price-sensitive consumers may benefit from exploring pricing variations across vendors to find the best deals, while retailers may face pressure to adjust their prices to remain competitive.
 
-## Third discussion point
+The stability of historical prices as a predictor of current prices also highlights potential concerns about price rigidity in certain product categories.
+This rigidity may limit the effectiveness of competition, as vendors might rely on historical price baselines rather than actively adjusting prices in response to market demand.
+Policymakers interested in promoting competition in the grocery sector could consider encouraging more frequent pricing updates or transparency in pricing strategies, particularly for staple items that heavily impact household budgets.
 
-## Weaknesses and next steps
+## Limitations of Data and Model
 
-Weaknesses and next steps should also be included.
+This analysis has several limitations that may impact the accuracy and comprehensiveness of our findings.
+First, the dataset is limited to major grocery chains, which may not capture price variations across smaller or regional stores that could offer different pricing structures.
+Additionally, the lack of granular product-level data limits our ability to analyze category-specific trends (e.g., dairy versus produce) or regional differences within the same vendor.
+
+The model’s focus on historical prices, vendor identity, and price differences also excludes potential external factors that may influence pricing, such as seasonal demand, promotions, or supply chain constraints.
+Including additional variables, such as the frequency of price adjustments or regional economic indicators, could enhance the model’s predictive power and provide a more comprehensive view of grocery pricing dynamics.
+
+## Future Research Direction
+
+Future research could build on this work by incorporating a wider range of variables, including product-specific attributes, seasonal factors, and promotional data, to capture a broader spectrum of influences on grocery prices.
+Exploring how demographic factors, such as income levels in different regions, correlate with price patterns across vendors could offer insights into the socio-economic impacts of grocery pricing on different communities.
+
+Increasing the model’s granularity to analyze price dynamics at the regional or product-category level would provide a deeper understanding of how specific items are priced within the same store or across regions.
+Additionally, expanding the analysis to include other sectors, such as online grocery prices or specialty stores, could provide a more holistic view of the Canadian grocery market.
+
+Incorporating machine learning techniques could also improve the model's adaptability to rapidly changing market conditions, allowing for more nuanced forecasts of pricing trends.
+This approach could be particularly useful in the face of ongoing challenges such as inflation and supply chain disruptions, offering a valuable tool for policymakers and consumers interested in understanding and mitigating grocery costs in Canada.
 
@@
 
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
 \newpage

### 2024-11-13 paper/paper.tex
Added visuals

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={My title},
  pdfauthor={First author; Another author},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{My title\thanks{Code and data are available at:
\url{https://github.com/RohanAlexander/starter_folder}.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{My subtitle if needed}
\author{First author \and Another author}
\date{November 13, 2024}

\begin{document}
\maketitle
\begin{abstract}
First sentence. Second sentence. Third sentence. Fourth sentence.
\end{abstract}

\section{Introduction}\label{introduction}

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as
follows. Section~\ref{sec-data}\ldots.

\section{Data}\label{sec-data}

\subsection{Overview}\label{overview}

For this study, we utilize data from Project Hammer, a Canadian
initiative designed to monitor grocery prices across major retailers in
Canada. The Project Hammer dataset was collected from eight prominent
Canadian grocery vendors---Voila, T\&T, Loblaws, No Frills, Metro,
Galleria, Walmart, and Save-On-Foods---between February 28, 2024, and
the latest available data load in the database
(\textbf{citeProjectHammerData?}). This dataset enables us to
investigate price fluctuations, identify pricing trends, and explore
potential competitive or collusive patterns within the Canadian grocery
sector. Project Hammer aims to support regulatory efforts by providing
data on price transparency, making it suitable for academic research and
potential legal analysis. Following data processing and cleaning
procedures outlined in Section 3.2, the dataset was structured to allow
an analysis of historical price changes and vendor-specific price
patterns. The dataset contains 1,996,969 rows and 5 columns, with
variables capturing product (product name), vendor\_name (name of the
grocery retailer), price\_current (current product price at the time of
recording), price\_old (previous recorded price for the product), and
price\_difference (difference between the current and old prices). Price
data is recorded in Canadian dollars and captures a broad range of
grocery items from various categories, including fresh produce, dairy,
pantry staples, and household items. To ensure the quality and
consistency of the data, we focused on removing missing values and
imputing outlier prices when necessary. For this analysis, we excluded
entries with extreme price fluctuations beyond three standard
deviations, as these could represent temporary discounts or data
recording anomalies. This cleaning process provides a balanced and
accurate reflection of grocery pricing, allowing for robust statistical
analysis on vendor-specific trends and average price comparisons across
product types. By narrowing the scope to recent data, Project Hammer's
dataset captures real-time fluctuations in grocery pricing, providing
timely insights into potential market dynamics and consumer price
impacts. The analysis allows for future investigation of temporal price
trends, vendor-based price differences, and, more broadly, consumer
affordability in a market facing rising cost-of-living pressures.

\subsection{Measurement}\label{measurement}

Some paragraphs about how we go from a phenomena in the world to an
entry in the dataset.

\subsection{Outcome variables}\label{outcome-variables}

Add graphs, tables and text. Use sub-sub-headings for each outcome
variable or update the subheading to be singular.

Some of our data is of penguins (Figure~\ref{fig-bills}), from Horst,
Hill, and Gorman (2020).

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-bills-1.pdf}

}

\caption{\label{fig-bills}Bills of penguins}

\end{figure}%

Talk more about it.

And also planes (Figure~\ref{fig-planes}). (You can change the height
and width, but don't worry about doing that until you have finished
every other aspect of the paper - Quarto will try to make it look nice
and the defaults usually work well once you have enough text.)

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-planes-1.pdf}

}

\caption{\label{fig-planes}Relationship between wing length and width}

\end{figure}%

Talk way more about it.

\subsection{Predictor variables}\label{predictor-variables}

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine
a few into one if they go together naturally.

\section{Model}\label{model}

Our modeling approach seeks to explore and quantify the relationship
between previous grocery prices, vendor identities, and price
differences in the current prices observed within Canadian grocery
stores. This analysis employs a Bayesian linear regression model
implemented via the \texttt{stan\ \_glm} function in the
\texttt{rstanarm} package to examine how factors such as historical
prices, vendor differences, and observed price changes impact current
prices for various grocery items.

In this model, \texttt{price\_current} serves as the response variable,
while \texttt{price\_old}, \texttt{vendor\_name}, and
\texttt{price\_difference} act as predictor variables. The linear
regression model assumes a Gaussian distribution for the response
variable \texttt{price\_current}, allowing for a straightforward
interpretation of the estimated parameters.

\subsection{Model set-up}\label{model-set-up}

The model includes the following predictor variables:

\begin{itemize}
\item
  Previous Price (\texttt{price\_old}): The price of the product in a
  previous time period.
\item
  Vendor (\texttt{vendor\_name}): A categorical variable representing
  the grocery store chain selling the product, capturing vendor-specific
  pricing differences.
\item
  Price Difference (\texttt{price\_difference}): The difference between
  the current and previous price, which may indicate market or
  vendor-specific pricing adjustments.
\end{itemize}

The model can be represented mathematically as follows:

\begin{align*}
    y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
    \mu_i &= \beta_0 + \beta_1 \cdot \text{Previous Price}_i + \beta_2 \cdot \text{Vendor}_i + \beta_3 \cdot \text{Price Difference}_i \\
    \epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

\textbf{Where:}

\begin{itemize}
    \item $\beta_0$ is the intercept term, representing the baseline estimage of `price_current`
    \item $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients representing the effects of `price_old`, `vendor_name` and `price_difference` on `price_current`.
    \item $\sigma^2$represents the variance of the error term, capturing unexplained variability in current prices.
\end{itemize}

The model is executed in \texttt{R} (R Core Team 2023) using the
\texttt{rstanarm} package (Goodrich et al. 2022), with priors set to
regularize the estimates and prevent overfitting. Specifically, we use a
normal prior for the coefficients, centered at zero with moderate
variance, to ensure stable estimates without overly restrictive
assumptions.

\subsubsection{Model justification}\label{model-justification}

The selection of a linear regression model for this analysis is based on
the continuous nature of \texttt{price\_current}, which allows us to
quantify how previous pricing, vendor, and price changes influence
current pricing. Existing economic theories suggest that previous prices
and historical pricing patterns are often predictive of current pricing,
especially in competitive retail markets where price sensitivity and
vendor-specific strategies play significant roles. The inclusion of
\texttt{price\_old} as a predictor aligns with time-series economic
models where past data points inform future values.

Additionally, vendor differences, represented by \texttt{vendor\_name},
capture potential competitive dynamics in the Canadian grocery market.
Vendors may adopt distinct pricing strategies or respond differently to
market conditions, and the inclusion of this categorical variable
enables an analysis of these patterns across various major grocery
chains. The \texttt{price\_difference} variable, which represents recent
price changes, provides an understanding of market adjustments and
reflects fluctuations potentially influenced by external factors, such
as supply chain constraints or inflation.

The Bayesian approach was selected for its flexibility in incorporating
prior information and its robustness in handling uncertainty within
small or moderate sample sizes. Furthermore, Bayesian methods allow us
to capture the posterior distributions of the parameters, which can be
used to evaluate the strength and credibility of each predictor's effect
on current prices. This approach also aligns with the objectives of
Project Hammer by facilitating a transparent and probabilistically
rigorous analysis of grocery pricing trends.

In summary, this model helps explain the primary factors of current
grocery prices in Canadian stores, accounting for both historical prices
and vendor effects. By utilizing Bayesian framework, this analysis
provides a clear quantification of the influence of each predictor while
allowing for robust estimation and interpretability in the context of
economic and competitive pricing strategies.

\section{Results}\label{results}

The results of our analysis of the Project Hammer dataset explain
important findings about pricing trends across major Canadian grocery
vendors. By examining factors such as prior pricing
(\texttt{price\_old}), vendor identities (\texttt{vendor\_name}), and
recent price changes (\texttt{price\_difference}), we assess how these
variables affect current prices (\texttt{price\_current}). The following
sections summarize key findings, supported by visualizations that
highlight trends and relationships within the data.

\subsection{Vendor-Specific Price
Trends}\label{vendor-specific-price-trends}

We begin by examining average current prices (\texttt{price\_current})
across different grocery vendors to identify potential pricing
variations in the Canadian grocery market.

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\#\#\# Read data \#\#\#\#}
\CommentTok{\# Load the cleaned dataset}
\NormalTok{analysis\_data }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"C:/Users/eliza/Downloads/hammer{-}3{-}compressed/cleaned\_price\_changes.csv"}\NormalTok{, }\AttributeTok{show\_col\_types =} \ConstantTok{FALSE}\NormalTok{)}

\CommentTok{\# Generate plot for Average Current Price by Vendor}
\NormalTok{vendor\_price\_plot }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(vendor\_name) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{summarize}\NormalTok{(}\AttributeTok{avg\_price\_current =} \FunctionTok{mean}\NormalTok{(price\_current, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{reorder}\NormalTok{(vendor\_name, avg\_price\_current), }\AttributeTok{y =}\NormalTok{ avg\_price\_current)) }\SpecialCharTok{+}
  \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{stat =} \StringTok{"identity"}\NormalTok{, }\AttributeTok{fill =} \StringTok{"skyblue"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{coord\_flip}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}
    \AttributeTok{title =} \StringTok{"Average Current Price by Vendor"}\NormalTok{,}
    \AttributeTok{x =} \StringTok{"Vendor"}\NormalTok{,}
    \AttributeTok{y =} \StringTok{"Average Current Price (CAD)"}
\NormalTok{  ) }\SpecialCharTok{+}
  \FunctionTok{theme\_minimal}\NormalTok{()}

\CommentTok{\# Display the plot}
\NormalTok{vendor\_price\_plot}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-1.pdf}

Figure 1: Average Current Price by Vendor. This bar plot shows the
average price of products sold by each grocery vendor in Canadian
dollars (CAD). Variations in pricing suggest differences in market
positioning or pricing strategies among vendors, with some retailers
consistently offering higher or lower prices across product categories.

\subsection{Price Difference Analysis}\label{price-difference-analysis}

We further analyze \texttt{price\_difference}, which captures the change
between the current price (\texttt{price\_current}) and the previous
price (\texttt{price\_old}). This variable provides insights into recent
price adjustments across vendors.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Average price difference by vendor}
\NormalTok{price\_diff\_plot }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(vendor\_name) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{summarize}\NormalTok{(}\AttributeTok{avg\_price\_difference =} \FunctionTok{mean}\NormalTok{(price\_difference, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{reorder}\NormalTok{(vendor\_name, avg\_price\_difference), }\AttributeTok{y =}\NormalTok{ avg\_price\_difference)) }\SpecialCharTok{+}
  \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{stat =} \StringTok{"identity"}\NormalTok{, }\AttributeTok{fill =} \StringTok{"coral"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{coord\_flip}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}
    \AttributeTok{title =} \StringTok{"Average Price Difference by Vendor"}\NormalTok{,}
    \AttributeTok{x =} \StringTok{"Vendor"}\NormalTok{,}
    \AttributeTok{y =} \StringTok{"Average Price Difference (CAD)"}
\NormalTok{  ) }\SpecialCharTok{+}
  \FunctionTok{theme\_minimal}\NormalTok{()}
\NormalTok{price\_diff\_plot}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-5-1.pdf}

Figure 2: Average Price Difference by Vendor. This bar plot shows the
average price change per vendor, highlighting retailers with the most
significant pricing adjustments. Positive differences suggest recent
price increases, while negative differences indicate reductions,
reflecting possible discounting or competitive pricing strategies.

\subsection{Relationship Between Previous and Current
Prices}\label{relationship-between-previous-and-current-prices}

We assess the relationship between previous prices (\texttt{price\_old})
and current prices (\texttt{price\_current}) to understand price
consistency and potential price elasticity across products.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Scatter plot of previous price vs. current price}
\NormalTok{price\_relationship\_plot }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(analysis\_data, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ price\_old, }\AttributeTok{y =}\NormalTok{ price\_current)) }\SpecialCharTok{+}
  \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.4}\NormalTok{, }\AttributeTok{color =} \StringTok{"blue"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{geom\_smooth}\NormalTok{(}\AttributeTok{method =} \StringTok{"lm"}\NormalTok{, }\AttributeTok{color =} \StringTok{"red"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}
    \AttributeTok{title =} \StringTok{"Relationship Between Previous and Current Prices"}\NormalTok{,}
    \AttributeTok{x =} \StringTok{"Previous Price (CAD)"}\NormalTok{,}
    \AttributeTok{y =} \StringTok{"Current Price (CAD)"}
\NormalTok{  ) }\SpecialCharTok{+}
  \FunctionTok{theme\_minimal}\NormalTok{()}
\NormalTok{price\_relationship\_plot}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
`geom_smooth()` using formula = 'y ~ x'
\end{verbatim}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-6-1.pdf}

Figure 3: Relationship Between Previous and Current Prices. This scatter
plot with a linear regression line shows the correlation between
previous and current prices. A positive relationship would indicate that
higher previous prices predict higher current prices, implying potential
price stability or gradual price changes in the grocery market.

\subsection{Distribution of Price
Differences}\label{distribution-of-price-differences}

To further explore pricing dynamics, we examine the distribution of
\texttt{price\_difference}, capturing both price increases and decreases
across the dataset.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Calculate mean and standard deviation of price\_difference}
\NormalTok{mean\_diff }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(analysis\_data}\SpecialCharTok{$}\NormalTok{price\_difference, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
\NormalTok{sd\_diff }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(analysis\_data}\SpecialCharTok{$}\NormalTok{price\_difference, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}

\CommentTok{\# Filter out extreme values beyond 3 standard deviations}
\NormalTok{filtered\_data }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(price\_difference }\SpecialCharTok{\textgreater{}=}\NormalTok{ (mean\_diff }\SpecialCharTok{{-}} \DecValTok{3} \SpecialCharTok{*}\NormalTok{ sd\_diff),}
\NormalTok{         price\_difference }\SpecialCharTok{\textless{}=}\NormalTok{ (mean\_diff }\SpecialCharTok{+} \DecValTok{3} \SpecialCharTok{*}\NormalTok{ sd\_diff))}

\CommentTok{\# Plot histogram with filtered data}
\NormalTok{price\_difference\_hist }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(filtered\_data, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ price\_difference)) }\SpecialCharTok{+}
  \FunctionTok{geom\_histogram}\NormalTok{(}\AttributeTok{bins =} \DecValTok{30}\NormalTok{, }\AttributeTok{fill =} \StringTok{"lightgreen"}\NormalTok{, }\AttributeTok{color =} \StringTok{"black"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}
    \AttributeTok{title =} \StringTok{"Distribution of Price Differences (Filtered)"}\NormalTok{,}
    \AttributeTok{x =} \StringTok{"Price Difference (CAD)"}\NormalTok{,}
    \AttributeTok{y =} \StringTok{"Frequency"}
\NormalTok{  ) }\SpecialCharTok{+}
  \FunctionTok{theme\_minimal}\NormalTok{()}
\NormalTok{price\_difference\_hist}
\end{Highlighting}
\end{Shaded}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-7-1.pdf}

Figure 4: Distribution of Price Differences (Filtered). This histogram
illustrates the frequency of price changes across products, with extreme
values beyond three standard deviations removed for clarity. The
concentration around zero suggests that small price adjustments are most
common, while the presence of both negative and positive values
indicates instances of both price increases and decreases. This
distribution provides insight into the stability and variability of
pricing across vendors. \#\# Model Evaluation

To evaluate our model's performance, we assess the predictive accuracy
of the Bayesian linear regression model on the test data. The model's
Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) provide
insights into the model's ability to predict \texttt{price\_current}
based on \texttt{price\_old}, \texttt{vendor\_name}, and
\texttt{price\_difference}.

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\#\#\# Read data \#\#\#\#}
\CommentTok{\# Load the cleaned data with data.table for efficiency}
\NormalTok{analysis\_data }\OtherTok{\textless{}{-}} \FunctionTok{fread}\NormalTok{(}\StringTok{"C:/Users/eliza/Downloads/hammer{-}3{-}compressed/cleaned\_price\_changes.csv"}\NormalTok{)}

\CommentTok{\# Split data into training and testing sets}
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{853}\NormalTok{)}
\NormalTok{train\_indices }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\FunctionTok{seq\_len}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data)), }\AttributeTok{size =} \FloatTok{0.8} \SpecialCharTok{*} \FunctionTok{nrow}\NormalTok{(analysis\_data))}
\NormalTok{analysis\_data\_train }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data[train\_indices]}
\NormalTok{analysis\_data\_test }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data[}\SpecialCharTok{{-}}\NormalTok{train\_indices]}
\DocumentationTok{\#\#\# Model data \#\#\#\#}
\CommentTok{\# Define a Bayesian regression model to predict \textasciigrave{}price\_current\textasciigrave{}}

\DocumentationTok{\#\#\# Model data \#\#\#\#}
\CommentTok{\# Define a Bayesian regression model to predict \textasciigrave{}price\_current\textasciigrave{}}
\NormalTok{model }\OtherTok{\textless{}{-}} \FunctionTok{stan\_glm}\NormalTok{(}
  \AttributeTok{formula =}\NormalTok{ price\_current }\SpecialCharTok{\textasciitilde{}}\NormalTok{ price\_old }\SpecialCharTok{+} \FunctionTok{as.factor}\NormalTok{(vendor\_name) }\SpecialCharTok{+}\NormalTok{ price\_difference,}
  \AttributeTok{data =}\NormalTok{ analysis\_data\_train,}
  \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
  \AttributeTok{prior =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \FloatTok{2.5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
  \AttributeTok{prior\_intercept =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \FloatTok{2.5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
  \AttributeTok{prior\_aux =} \FunctionTok{exponential}\NormalTok{(}\AttributeTok{rate =} \DecValTok{1}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
  \AttributeTok{seed =} \DecValTok{853}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}

SAMPLING FOR MODEL 'continuous' NOW (CHAIN 1).
Chain 1: 
Chain 1: Gradient evaluation took 0.001214 seconds
Chain 1: 1000 transitions using 10 leapfrog steps per transition would take 12.14 seconds.
Chain 1: Adjust your expectations accordingly!
Chain 1: 
Chain 1: 
Chain 1: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 1: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 1: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 1: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 1: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 1: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 1: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 1: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 1: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 1: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 1: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 1: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 1: 
Chain 1:  Elapsed Time: 0.56 seconds (Warm-up)
Chain 1:                225.422 seconds (Sampling)
Chain 1:                225.982 seconds (Total)
Chain 1:

SAMPLING FOR MODEL 'continuous' NOW (CHAIN 2).
Chain 2: 
Chain 2: Gradient evaluation took 3.6e-05 seconds
Chain 2: 1000 transitions using 10 leapfrog steps per transition would take 0.36 seconds.
Chain 2: Adjust your expectations accordingly!
Chain 2: 
Chain 2: 
Chain 2: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 2: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 2: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 2: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 2: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 2: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 2: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 2: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 2: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 2: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 2: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 2: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 2: 
Chain 2:  Elapsed Time: 0.47 seconds (Warm-up)
Chain 2:                208.219 seconds (Sampling)
Chain 2:                208.689 seconds (Total)
Chain 2:

SAMPLING FOR MODEL 'continuous' NOW (CHAIN 3).
Chain 3: 
Chain 3: Gradient evaluation took 4.6e-05 seconds
Chain 3: 1000 transitions using 10 leapfrog steps per transition would take 0.46 seconds.
Chain 3: Adjust your expectations accordingly!
Chain 3: 
Chain 3: 
Chain 3: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 3: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 3: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 3: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 3: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 3: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 3: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 3: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 3: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 3: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 3: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 3: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 3: 
Chain 3:  Elapsed Time: 0.522 seconds (Warm-up)
Chain 3:                207.839 seconds (Sampling)
Chain 3:                208.361 seconds (Total)
Chain 3:

SAMPLING FOR MODEL 'continuous' NOW (CHAIN 4).
Chain 4: 
Chain 4: Gradient evaluation took 2.7e-05 seconds
Chain 4: 1000 transitions using 10 leapfrog steps per transition would take 0.27 seconds.
Chain 4: Adjust your expectations accordingly!
Chain 4: 
Chain 4: 
Chain 4: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 4: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 4: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 4: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 4: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 4: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 4: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 4: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 4: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 4: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 4: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 4: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 4: 
Chain 4:  Elapsed Time: 0.31 seconds (Warm-up)
Chain 4:                210.47 seconds (Sampling)
Chain 4:                210.78 seconds (Total)
Chain 4: 
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Define the chunk size (adjust based on available memory)}
\NormalTok{chunk\_size }\OtherTok{\textless{}{-}} \DecValTok{50000}  
\NormalTok{num\_chunks }\OtherTok{\textless{}{-}} \FunctionTok{ceiling}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data\_test) }\SpecialCharTok{/}\NormalTok{ chunk\_size)}

\CommentTok{\# Initialize a vector to store predictions}
\NormalTok{predictions }\OtherTok{\textless{}{-}} \FunctionTok{vector}\NormalTok{(}\StringTok{"numeric"}\NormalTok{, }\FunctionTok{nrow}\NormalTok{(analysis\_data\_test))}

\CommentTok{\# Loop through each chunk and generate predictions}
\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{num\_chunks) \{}
\NormalTok{  start }\OtherTok{\textless{}{-}}\NormalTok{ ((i }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ chunk\_size) }\SpecialCharTok{+} \DecValTok{1}
\NormalTok{  end }\OtherTok{\textless{}{-}} \FunctionTok{min}\NormalTok{(i }\SpecialCharTok{*}\NormalTok{ chunk\_size, }\FunctionTok{nrow}\NormalTok{(analysis\_data\_test))}

  \CommentTok{\# Generate predictions for the current chunk}
\NormalTok{  predictions[start}\SpecialCharTok{:}\NormalTok{end] }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(model, }\AttributeTok{newdata =}\NormalTok{ analysis\_data\_test[start}\SpecialCharTok{:}\NormalTok{end, ])}
\NormalTok{\}}

\CommentTok{\# Calculate MAE and RMSE for model evaluation}
\NormalTok{mae }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{abs}\NormalTok{(predictions }\SpecialCharTok{{-}}\NormalTok{ analysis\_data\_test}\SpecialCharTok{$}\NormalTok{price\_current))}
\NormalTok{rmse }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(}\FunctionTok{mean}\NormalTok{((predictions }\SpecialCharTok{{-}}\NormalTok{ analysis\_data\_test}\SpecialCharTok{$}\NormalTok{price\_current)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{))}

\NormalTok{evaluation\_metrics }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
  \AttributeTok{Metric =} \FunctionTok{c}\NormalTok{(}\StringTok{"Mean Absolute Error (MAE)"}\NormalTok{, }\StringTok{"Root Mean Squared Error (RMSE)"}\NormalTok{),}
  \AttributeTok{Value =} \FunctionTok{c}\NormalTok{(}\FunctionTok{round}\NormalTok{(mae, }\DecValTok{2}\NormalTok{), }\FunctionTok{round}\NormalTok{(rmse, }\DecValTok{2}\NormalTok{))}
\NormalTok{)}

\CommentTok{\# Display the results table}
\FunctionTok{kable}\NormalTok{(evaluation\_metrics, }\AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"Evaluation Metric"}\NormalTok{, }\StringTok{"Value"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}lr@{}}
\toprule\noalign{}
Evaluation Metric & Value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mean Absolute Error (MAE) & 6.90 \\
Root Mean Squared Error (RMSE) & 17.77 \\
\end{longtable}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{print}\NormalTok{(evaluation\_metrics)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
                          Metric Value
1      Mean Absolute Error (MAE)  6.90
2 Root Mean Squared Error (RMSE) 17.77
\end{verbatim}

Table 1: Model Evaluation Metrics. The table displays the MAE and RMSE
for the model, reflecting the average and squared prediction errors,
respectively. Lower values indicate better predictive accuracy,
suggesting that the model effectively captures the relationship between
previous prices, vendor identities, and price differences.

These results demonstrate that vendor identity and previous prices are
significant predictors of current prices in the Canadian grocery sector,
with implications for understanding competitive strategies and consumer
price sensitivity in the market.

\section{Discussion}\label{discussion}

\subsection{Interpretation of Grocery Pricing
Dynamics}\label{interpretation-of-grocery-pricing-dynamics}

This study investigates the primary factors influencing current grocery
prices in Canada by examining the Project Hammer dataset. Our Bayesian
regression model reveals that prior prices (\texttt{price\_old}), vendor
identity (\texttt{vendor\_name}), and recent price differences
(\texttt{price\_difference}) play significant roles in determining
current prices. The influence of previous prices indicates price
continuity across time, suggesting that grocery items tend to exhibit
relatively stable pricing patterns with minor adjustments.
Vendor-specific effects highlight the role of retailer pricing
strategies, with some vendors maintaining consistently higher or lower
prices. These vendor variations may reflect differences in market
positioning, operational costs, or competitive strategies, impacting
price consistency across the grocery sector.

Additionally, the presence of both positive and negative values in the
\texttt{price\_difference} variable illustrates the prevalence of both
price increases and decreases, underscoring the dynamic nature of
grocery pricing in response to external market forces. External factors,
such as supply chain disruptions, inflation, and seasonal demand
fluctuations, likely contribute to the observed variability. This model
highlights the complexity of pricing dynamics and the importance of
historical pricing and vendor-specific factors in shaping current
grocery prices across major Canadian retailers.

\subsection{Implications for Competition and Consumer
Costs}\label{implications-for-competition-and-consumer-costs}

The findings from Project Hammer have broader implications for both
competition in the grocery sector and consumer expenses. Vendor-specific
price differences suggest that some retailers adopt competitive pricing
strategies to attract cost-sensitive consumers, while others may
leverage brand loyalty or perceived quality to justify higher prices. In
this context, price-sensitive consumers may benefit from exploring
pricing variations across vendors to find the best deals, while
retailers may face pressure to adjust their prices to remain
competitive.

The stability of historical prices as a predictor of current prices also
highlights potential concerns about price rigidity in certain product
categories. This rigidity may limit the effectiveness of competition, as
vendors might rely on historical price baselines rather than actively
adjusting prices in response to market demand. Policymakers interested
in promoting competition in the grocery sector could consider
encouraging more frequent pricing updates or transparency in pricing
strategies, particularly for staple items that heavily impact household
budgets.

\subsection{Limitations of Data and
Model}\label{limitations-of-data-and-model}

This analysis has several limitations that may impact the accuracy and
comprehensiveness of our findings. First, the dataset is limited to
major grocery chains, which may not capture price variations across
smaller or regional stores that could offer different pricing
structures. Additionally, the lack of granular product-level data limits
our ability to analyze category-specific trends (e.g., dairy versus
produce) or regional differences within the same vendor.

The model's focus on historical prices, vendor identity, and price
differences also excludes potential external factors that may influence
pricing, such as seasonal demand, promotions, or supply chain
constraints. Including additional variables, such as the frequency of
price adjustments or regional economic indicators, could enhance the
model's predictive power and provide a more comprehensive view of
grocery pricing dynamics.

\subsection{Future Research Direction}\label{future-research-direction}

Future research could build on this work by incorporating a wider range
of variables, including product-specific attributes, seasonal factors,
and promotional data, to capture a broader spectrum of influences on
grocery prices. Exploring how demographic factors, such as income levels
in different regions, correlate with price patterns across vendors could
offer insights into the socio-economic impacts of grocery pricing on
different communities.

Increasing the model's granularity to analyze price dynamics at the
regional or product-category level would provide a deeper understanding
of how specific items are priced within the same store or across
regions. Additionally, expanding the analysis to include other sectors,
such as online grocery prices or specialty stores, could provide a more
holistic view of the Canadian grocery market.

Incorporating machine learning techniques could also improve the model's
adaptability to rapidly changing market conditions, allowing for more
nuanced forecasts of pricing trends. This approach could be particularly
useful in the face of ongoing challenges such as inflation and supply
chain disruptions, offering a valuable tool for policymakers and
consumers interested in understanding and mitigating grocery costs in
Canada.

\newpage

\appendix

\section*{Appendix}\label{appendix}
\addcontentsline{toc}{section}{Appendix}

\section{Additional data details}\label{additional-data-details}

\section{Model details}\label{sec-model-details}

\subsection{Posterior predictive
check}\label{posterior-predictive-check}

In \textbf{?@fig-ppcheckandposteriorvsprior-1} we implement a posterior
predictive check. This shows\ldots{}

In \textbf{?@fig-ppcheckandposteriorvsprior-2} we compare the posterior
with the prior. This shows\ldots{}

\newpage

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-rstanarm}
Goodrich, Ben, Jonah Gabry, Imad Ali, and Sam Brilleman. 2022.
{``{rstanarm: {Bayesian} applied regression modeling via {Stan}}.''}
\url{https://mc-stan.org/rstanarm/}.

\bibitem[\citeproctext]{ref-palmerpenguins}
Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020.
\emph{{palmerpenguins: Palmer Archipelago (Antarctica) penguin data}}.
\url{https://doi.org/10.5281/zenodo.3960218}.

\bibitem[\citeproctext]{ref-citeR}
R Core Team. 2023. \emph{{R: A Language and Environment for Statistical
Computing}}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\end{CSLReferences}

\end{document}

### 2024-11-14 paper/paper.qmd
Updated code chunk 8

@@
 For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada.
-The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [@citeProjectHammerData].
+The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [ citeProjectHammerData].
 This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.
@@
 \begin{itemize}
-    \item $\beta_0$ is the intercept term, representing the baseline estimage of `price_current`
-    \item $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients representing the effects of `price_old`, `vendor_name` and `price_difference` on `price_current`.
+    \item $\beta_0$ is the intercept term, representing the baseline estimate of $price_{current}$
+    \item $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients representing the effects of $price_{old}$, $vendor_{name}$ and $price_{difference}$ on $price_{current}$.
     \item $\sigma^2$represents the variance of the error term, capturing unexplained variability in current prices.
@@
 
-#### Read data ####
-# Load the cleaned data with data.table for efficiency
+#### Read Data ####
+# Load only a sample of the data to reduce memory usage (e.g., 10% of the dataset)
+set.seed(853)
+
 analysis_data <- fread("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
+analysis_data <- analysis_data[sample(.N, size = 0.1 * .N)]
 
 # Split data into training and testing sets
-set.seed(853)
 train_indices <- sample(seq_len(nrow(analysis_data)), size = 0.8 * nrow(analysis_data))
@@
 analysis_data_test <- analysis_data[-train_indices]
-### Model data ####
-# Define a Bayesian regression model to predict `price_current`
 
-### Model data ####
-# Define a Bayesian regression model to predict `price_current`
+#### Model Data ####
+# Define a Bayesian regression model to predict `price_current` with fewer iterations and lighter priors
+
 model <- stan_glm(
@@
   family = gaussian(),
-  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-  prior_aux = exponential(rate = 1, autoscale = TRUE),
+  prior = normal(location = 0, scale = 5, autoscale = TRUE),      # Slightly broader prior
+  prior_intercept = normal(location = 0, scale = 5, autoscale = TRUE),
+  prior_aux = exponential(rate = 0.5, autoscale = TRUE),          # Less informative prior
+  iter = 500,                                                     # Reduce iterations
+  chains = 2,                                                     # Fewer chains for faster computation
   seed = 853
@@
 
-# Define the chunk size (adjust based on available memory)
-chunk_size <- 50000  
+#### Predict in Chunks ####
+# Define chunk size based on available memory
+chunk_size <- 20000  # Adjusted smaller chunk size for efficiency
 num_chunks <- ceiling(nrow(analysis_data_test) / chunk_size)
@@
 # Initialize a vector to store predictions
-predictions <- vector("numeric", nrow(analysis_data_test))
+predictions <- numeric(nrow(analysis_data_test))
 
-# Loop through each chunk and generate predictions
-for (i in 1:num_chunks) {
+# Loop through each chunk to generate predictions
+for (i in seq_len(num_chunks)) {
   start <- ((i - 1) * chunk_size) + 1
@@
 
-# Calculate MAE and RMSE for model evaluation
+#### Model Evaluation Metrics ####
+# Calculate Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE)
 mae <- mean(abs(predictions - analysis_data_test$price_current))
@@
 
-evaluation_metrics <- data.frame(
+# Store results in a data frame
+evaluation_metrics <- data.table(
   Metric = c("Mean Absolute Error (MAE)", "Root Mean Squared Error (RMSE)"),
@@
 # Display the results table
-kable(evaluation_metrics, col.names = c("Evaluation Metric", "Value"))
-
 print(evaluation_metrics)
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
 \newpage

### 2024-11-14 paper/paper.tex
Updated code chunk 8

@@
 \author{First author \and Another author}
-\date{November 13, 2024}
+\date{November 14, 2024}
 
@@
 Galleria, Walmart, and Save-On-Foods---between February 28, 2024, and
-the latest available data load in the database
-(\textbf{citeProjectHammerData?}). This dataset enables us to
-investigate price fluctuations, identify pricing trends, and explore
-potential competitive or collusive patterns within the Canadian grocery
-sector. Project Hammer aims to support regulatory efforts by providing
-data on price transparency, making it suitable for academic research and
-potential legal analysis. Following data processing and cleaning
-procedures outlined in Section 3.2, the dataset was structured to allow
-an analysis of historical price changes and vendor-specific price
-patterns. The dataset contains 1,996,969 rows and 5 columns, with
-variables capturing product (product name), vendor\_name (name of the
-grocery retailer), price\_current (current product price at the time of
-recording), price\_old (previous recorded price for the product), and
+the latest available data load in the database {[}
+citeProjectHammerData{]}. This dataset enables us to investigate price
+fluctuations, identify pricing trends, and explore potential competitive
+or collusive patterns within the Canadian grocery sector. Project Hammer
+aims to support regulatory efforts by providing data on price
+transparency, making it suitable for academic research and potential
+legal analysis. Following data processing and cleaning procedures
+outlined in Section 3.2, the dataset was structured to allow an analysis
+of historical price changes and vendor-specific price patterns. The
+dataset contains 1,996,969 rows and 5 columns, with variables capturing
+product (product name), vendor\_name (name of the grocery retailer),
+price\_current (current product price at the time of recording),
+price\_old (previous recorded price for the product), and
 price\_difference (difference between the current and old prices). Price
@@
 \begin{Highlighting}[]
-\DocumentationTok{\#\#\#\# Read data \#\#\#\#}
-\CommentTok{\# Load the cleaned data with data.table for efficiency}
+\DocumentationTok{\#\#\#\# Read Data \#\#\#\#}
+\CommentTok{\# Load only a sample of the data to reduce memory usage (e.g., 10\% of the dataset)}
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{853}\NormalTok{)}
+
 \NormalTok{analysis\_data }\OtherTok{\textless{}{-}} \FunctionTok{fread}\NormalTok{(}\StringTok{"C:/Users/eliza/Downloads/hammer{-}3{-}compressed/cleaned\_price\_changes.csv"}\NormalTok{)}
+\NormalTok{analysis\_data }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data[}\FunctionTok{sample}\NormalTok{(.N, }\AttributeTok{size =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ .N)]}
 
 \CommentTok{\# Split data into training and testing sets}
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{853}\NormalTok{)}
 \NormalTok{train\_indices }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\FunctionTok{seq\_len}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data)), }\AttributeTok{size =} \FloatTok{0.8} \SpecialCharTok{*} \FunctionTok{nrow}\NormalTok{(analysis\_data))}
@@
 \NormalTok{analysis\_data\_test }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data[}\SpecialCharTok{{-}}\NormalTok{train\_indices]}
-\DocumentationTok{\#\#\# Model data \#\#\#\#}
-\CommentTok{\# Define a Bayesian regression model to predict \textasciigrave{}price\_current\textasciigrave{}}
 
-\DocumentationTok{\#\#\# Model data \#\#\#\#}
-\CommentTok{\# Define a Bayesian regression model to predict \textasciigrave{}price\_current\textasciigrave{}}
+\DocumentationTok{\#\#\#\# Model Data \#\#\#\#}
+\CommentTok{\# Define a Bayesian regression model to predict \textasciigrave{}price\_current\textasciigrave{} with fewer iterations and lighter priors}
+\FunctionTok{library}\NormalTok{(rstanarm)}
 \NormalTok{model }\OtherTok{\textless{}{-}} \FunctionTok{stan\_glm}\NormalTok{(}
@@
   \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
-  \AttributeTok{prior =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \FloatTok{2.5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
-  \AttributeTok{prior\_intercept =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \FloatTok{2.5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
-  \AttributeTok{prior\_aux =} \FunctionTok{exponential}\NormalTok{(}\AttributeTok{rate =} \DecValTok{1}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
+  \AttributeTok{prior =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \DecValTok{5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),      }\CommentTok{\# Slightly broader prior}
+  \AttributeTok{prior\_intercept =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \DecValTok{5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
+  \AttributeTok{prior\_aux =} \FunctionTok{exponential}\NormalTok{(}\AttributeTok{rate =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),          }\CommentTok{\# Less informative prior}
+  \AttributeTok{iter =} \DecValTok{500}\NormalTok{,                                                     }\CommentTok{\# Reduce iterations}
+  \AttributeTok{chains =} \DecValTok{2}\NormalTok{,                                                     }\CommentTok{\# Fewer chains for faster computation}
   \AttributeTok{seed =} \DecValTok{853}
@@
 Chain 1: 
-Chain 1: Gradient evaluation took 0.001214 seconds
-Chain 1: 1000 transitions using 10 leapfrog steps per transition would take 12.14 seconds.
+Chain 1: Gradient evaluation took 0.000105 seconds
+Chain 1: 1000 transitions using 10 leapfrog steps per transition would take 1.05 seconds.
 Chain 1: Adjust your expectations accordingly!
@@
 Chain 1: 
-Chain 1: Iteration:    1 / 2000 [  0%]  (Warmup)
-Chain 1: Iteration:  200 / 2000 [ 10%]  (Warmup)
-Chain 1: Iteration:  400 / 2000 [ 20%]  (Warmup)
-Chain 1: Iteration:  600 / 2000 [ 30%]  (Warmup)
-Chain 1: Iteration:  800 / 2000 [ 40%]  (Warmup)
-Chain 1: Iteration: 1000 / 2000 [ 50%]  (Warmup)
-Chain 1: Iteration: 1001 / 2000 [ 50%]  (Sampling)
-Chain 1: Iteration: 1200 / 2000 [ 60%]  (Sampling)
-Chain 1: Iteration: 1400 / 2000 [ 70%]  (Sampling)
-Chain 1: Iteration: 1600 / 2000 [ 80%]  (Sampling)
-Chain 1: Iteration: 1800 / 2000 [ 90%]  (Sampling)
-Chain 1: Iteration: 2000 / 2000 [100%]  (Sampling)
+Chain 1: Iteration:   1 / 500 [  0%]  (Warmup)
+Chain 1: Iteration:  50 / 500 [ 10%]  (Warmup)
+Chain 1: Iteration: 100 / 500 [ 20%]  (Warmup)
+Chain 1: Iteration: 150 / 500 [ 30%]  (Warmup)
+Chain 1: Iteration: 200 / 500 [ 40%]  (Warmup)
+Chain 1: Iteration: 250 / 500 [ 50%]  (Warmup)
+Chain 1: Iteration: 251 / 500 [ 50%]  (Sampling)
+Chain 1: Iteration: 300 / 500 [ 60%]  (Sampling)
+Chain 1: Iteration: 350 / 500 [ 70%]  (Sampling)
+Chain 1: Iteration: 400 / 500 [ 80%]  (Sampling)
+Chain 1: Iteration: 450 / 500 [ 90%]  (Sampling)
+Chain 1: Iteration: 500 / 500 [100%]  (Sampling)
 Chain 1: 
-Chain 1:  Elapsed Time: 0.56 seconds (Warm-up)
-Chain 1:                225.422 seconds (Sampling)
-Chain 1:                225.982 seconds (Total)
+Chain 1:  Elapsed Time: 0.056 seconds (Warm-up)
+Chain 1:                3.902 seconds (Sampling)
+Chain 1:                3.958 seconds (Total)
 Chain 1:
@@
 Chain 2: 
-Chain 2: Gradient evaluation took 3.6e-05 seconds
-Chain 2: 1000 transitions using 10 leapfrog steps per transition would take 0.36 seconds.
+Chain 2: Gradient evaluation took 2.5e-05 seconds
+Chain 2: 1000 transitions using 10 leapfrog steps per transition would take 0.25 seconds.
 Chain 2: Adjust your expectations accordingly!
@@
 Chain 2: 
-Chain 2: Iteration:    1 / 2000 [  0%]  (Warmup)
-Chain 2: Iteration:  200 / 2000 [ 10%]  (Warmup)
-Chain 2: Iteration:  400 / 2000 [ 20%]  (Warmup)
-Chain 2: Iteration:  600 / 2000 [ 30%]  (Warmup)
-Chain 2: Iteration:  800 / 2000 [ 40%]  (Warmup)
-Chain 2: Iteration: 1000 / 2000 [ 50%]  (Warmup)
-Chain 2: Iteration: 1001 / 2000 [ 50%]  (Sampling)
-Chain 2: Iteration: 1200 / 2000 [ 60%]  (Sampling)
-Chain 2: Iteration: 1400 / 2000 [ 70%]  (Sampling)
-Chain 2: Iteration: 1600 / 2000 [ 80%]  (Sampling)
-Chain 2: Iteration: 1800 / 2000 [ 90%]  (Sampling)
-Chain 2: Iteration: 2000 / 2000 [100%]  (Sampling)
+Chain 2: Iteration:   1 / 500 [  0%]  (Warmup)
+Chain 2: Iteration:  50 / 500 [ 10%]  (Warmup)
+Chain 2: Iteration: 100 / 500 [ 20%]  (Warmup)
+Chain 2: Iteration: 150 / 500 [ 30%]  (Warmup)
+Chain 2: Iteration: 200 / 500 [ 40%]  (Warmup)
+Chain 2: Iteration: 250 / 500 [ 50%]  (Warmup)
+Chain 2: Iteration: 251 / 500 [ 50%]  (Sampling)
+Chain 2: Iteration: 300 / 500 [ 60%]  (Sampling)
+Chain 2: Iteration: 350 / 500 [ 70%]  (Sampling)
+Chain 2: Iteration: 400 / 500 [ 80%]  (Sampling)
+Chain 2: Iteration: 450 / 500 [ 90%]  (Sampling)
+Chain 2: Iteration: 500 / 500 [100%]  (Sampling)
 Chain 2: 
-Chain 2:  Elapsed Time: 0.47 seconds (Warm-up)
-Chain 2:                208.219 seconds (Sampling)
-Chain 2:                208.689 seconds (Total)
+Chain 2:  Elapsed Time: 0.036 seconds (Warm-up)
+Chain 2:                3.936 seconds (Sampling)
+Chain 2:                3.972 seconds (Total)
 Chain 2: 
+\end{verbatim}
 
-SAMPLING FOR MODEL 'continuous' NOW (CHAIN 3).
-Chain 3: 
-Chain 3: Gradient evaluation took 4.6e-05 seconds
-Chain 3: 1000 transitions using 10 leapfrog steps per transition would take 0.46 seconds.
-Chain 3: Adjust your expectations accordingly!
-Chain 3: 
-Chain 3: 
-Chain 3: Iteration:    1 / 2000 [  0%]  (Warmup)
-Chain 3: Iteration:  200 / 2000 [ 10%]  (Warmup)
-Chain 3: Iteration:  400 / 2000 [ 20%]  (Warmup)
-Chain 3: Iteration:  600 / 2000 [ 30%]  (Warmup)
-Chain 3: Iteration:  800 / 2000 [ 40%]  (Warmup)
-Chain 3: Iteration: 1000 / 2000 [ 50%]  (Warmup)
-Chain 3: Iteration: 1001 / 2000 [ 50%]  (Sampling)
-Chain 3: Iteration: 1200 / 2000 [ 60%]  (Sampling)
-Chain 3: Iteration: 1400 / 2000 [ 70%]  (Sampling)
-Chain 3: Iteration: 1600 / 2000 [ 80%]  (Sampling)
-Chain 3: Iteration: 1800 / 2000 [ 90%]  (Sampling)
-Chain 3: Iteration: 2000 / 2000 [100%]  (Sampling)
-Chain 3: 
-Chain 3:  Elapsed Time: 0.522 seconds (Warm-up)
-Chain 3:                207.839 seconds (Sampling)
-Chain 3:                208.361 seconds (Total)
-Chain 3:
+\begin{verbatim}
+Warning: There were 1 chains where the estimated Bayesian Fraction of Missing Information was low. See
+https://mc-stan.org/misc/warnings.html#bfmi-low
+\end{verbatim}
 
-SAMPLING FOR MODEL 'continuous' NOW (CHAIN 4).
-Chain 4: 
-Chain 4: Gradient evaluation took 2.7e-05 seconds
-Chain 4: 1000 transitions using 10 leapfrog steps per transition would take 0.27 seconds.
-Chain 4: Adjust your expectations accordingly!
-Chain 4: 
-Chain 4: 
-Chain 4: Iteration:    1 / 2000 [  0%]  (Warmup)
-Chain 4: Iteration:  200 / 2000 [ 10%]  (Warmup)
-Chain 4: Iteration:  400 / 2000 [ 20%]  (Warmup)
-Chain 4: Iteration:  600 / 2000 [ 30%]  (Warmup)
-Chain 4: Iteration:  800 / 2000 [ 40%]  (Warmup)
-Chain 4: Iteration: 1000 / 2000 [ 50%]  (Warmup)
-Chain 4: Iteration: 1001 / 2000 [ 50%]  (Sampling)
-Chain 4: Iteration: 1200 / 2000 [ 60%]  (Sampling)
-Chain 4: Iteration: 1400 / 2000 [ 70%]  (Sampling)
-Chain 4: Iteration: 1600 / 2000 [ 80%]  (Sampling)
-Chain 4: Iteration: 1800 / 2000 [ 90%]  (Sampling)
-Chain 4: Iteration: 2000 / 2000 [100%]  (Sampling)
-Chain 4: 
-Chain 4:  Elapsed Time: 0.31 seconds (Warm-up)
-Chain 4:                210.47 seconds (Sampling)
-Chain 4:                210.78 seconds (Total)
-Chain 4: 
+\begin{verbatim}
+Warning: Examine the pairs() plot to diagnose sampling problems
+\end{verbatim}
+
+\begin{verbatim}
+Warning: The largest R-hat is 1.6, indicating chains have not mixed.
+Running the chains for more iterations may help. See
+https://mc-stan.org/misc/warnings.html#r-hat
+\end{verbatim}
+
+\begin{verbatim}
+Warning: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.
+Running the chains for more iterations may help. See
+https://mc-stan.org/misc/warnings.html#bulk-ess
+\end{verbatim}
+
+\begin{verbatim}
+Warning: Tail Effective Samples Size (ESS) is too low, indicating posterior variances and tail quantiles may be unreliable.
+Running the chains for more iterations may help. See
+https://mc-stan.org/misc/warnings.html#tail-ess
+\end{verbatim}
+
+\begin{verbatim}
+Warning: Markov chains did not converge! Do not analyze results!
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\CommentTok{\# Define the chunk size (adjust based on available memory)}
-\NormalTok{chunk\_size }\OtherTok{\textless{}{-}} \DecValTok{50000}  
+\DocumentationTok{\#\#\#\# Predict in Chunks \#\#\#\#}
+\CommentTok{\# Define chunk size based on available memory}
+\NormalTok{chunk\_size }\OtherTok{\textless{}{-}} \DecValTok{20000}  \CommentTok{\# Adjusted smaller chunk size for efficiency}
 \NormalTok{num\_chunks }\OtherTok{\textless{}{-}} \FunctionTok{ceiling}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data\_test) }\SpecialCharTok{/}\NormalTok{ chunk\_size)}
@@
 \CommentTok{\# Initialize a vector to store predictions}
-\NormalTok{predictions }\OtherTok{\textless{}{-}} \FunctionTok{vector}\NormalTok{(}\StringTok{"numeric"}\NormalTok{, }\FunctionTok{nrow}\NormalTok{(analysis\_data\_test))}
+\NormalTok{predictions }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data\_test))}
 
-\CommentTok{\# Loop through each chunk and generate predictions}
-\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{num\_chunks) \{}
+\CommentTok{\# Loop through each chunk to generate predictions}
+\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \FunctionTok{seq\_len}\NormalTok{(num\_chunks)) \{}
 \NormalTok{  start }\OtherTok{\textless{}{-}}\NormalTok{ ((i }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ chunk\_size) }\SpecialCharTok{+} \DecValTok{1}
@@
 
-\CommentTok{\# Calculate MAE and RMSE for model evaluation}
+\DocumentationTok{\#\#\#\# Model Evaluation Metrics \#\#\#\#}
+\CommentTok{\# Calculate Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE)}
 \NormalTok{mae }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{abs}\NormalTok{(predictions }\SpecialCharTok{{-}}\NormalTok{ analysis\_data\_test}\SpecialCharTok{$}\NormalTok{price\_current))}
@@
 
-\NormalTok{evaluation\_metrics }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
+\CommentTok{\# Store results in a data frame}
+\NormalTok{evaluation\_metrics }\OtherTok{\textless{}{-}} \FunctionTok{data.table}\NormalTok{(}
   \AttributeTok{Metric =} \FunctionTok{c}\NormalTok{(}\StringTok{"Mean Absolute Error (MAE)"}\NormalTok{, }\StringTok{"Root Mean Squared Error (RMSE)"}\NormalTok{),}
@@
 \CommentTok{\# Display the results table}
-\FunctionTok{kable}\NormalTok{(evaluation\_metrics, }\AttributeTok{col.names =} \FunctionTok{c}\NormalTok{(}\StringTok{"Evaluation Metric"}\NormalTok{, }\StringTok{"Value"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{longtable}[]{@{}lr@{}}
-\toprule\noalign{}
-Evaluation Metric & Value \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Mean Absolute Error (MAE) & 6.90 \\
-Root Mean Squared Error (RMSE) & 17.77 \\
-\end{longtable}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \FunctionTok{print}\NormalTok{(evaluation\_metrics)}
@@
                            Metric Value
-1      Mean Absolute Error (MAE)  6.90
-2 Root Mean Squared Error (RMSE) 17.77
+                           <char> <num>
+1:      Mean Absolute Error (MAE)  6.74
+2: Root Mean Squared Error (RMSE) 17.98
 \end{verbatim}
@@
 
-In \textbf{?@fig-ppcheckandposteriorvsprior-1} we implement a posterior
-predictive check. This shows\ldots{}
-
-In \textbf{?@fig-ppcheckandposteriorvsprior-2} we compare the posterior
-with the prior. This shows\ldots{}
-
 \newpage

### 2024-11-14 paper/paper.qmd
Added fig-label

@@
 
-Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.
-
 ## Outcome variables
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
-
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
-
-Talk way more about it.
-
 ## Predictor variables
-
-Add graphs, tables and text.
-
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
 
@@
 ```{r}
+#| label: fig-avg-curr-price
+#| fig-cap: Average Current Price by Vendor.
+This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD).
+Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.
+#| echo: false
+#| warning: false
+#| message: false
+
 #### Read data ####
@@
 
-Figure 1: Average Current Price by Vendor.
-This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD).
-Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.
-
 ## Price Difference Analysis
@@
 ```{r}
+#| label: fig-avg-price-diff
+#| fig-cap: Average Price Difference by Vendor.
+This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments.
+Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.
+#| echo: false
+#| warning: false
+#| message: false
+
 # Average price difference by vendor
@@
 
-Figure 2: Average Price Difference by Vendor.
-This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments.
-Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.
-
 ## Relationship Between Previous and Current Prices
@@
 ```{r}
+#| label: fig-past-curr
+#| fig-cap: Relationship Between Previous and Current Prices.
+This scatter plot with a linear regression line shows the correlation between previous and current prices.
+A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
+#| echo: false
+#| warning: false
+#| message: false
+
 # Scatter plot of previous price vs. current price
@@
 ```
-Figure 3: Relationship Between Previous and Current Prices.
-This scatter plot with a linear regression line shows the correlation between previous and current prices.
-A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
 
@@
 ```{r}
+#| label: fig-dist-price
+#| fig-cap: Relationship Between Previous and Current Prices.
+This scatter plot with a linear regression line shows the correlation between previous and current prices.
+A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
+#| echo: false
+#| warning: false
+#| message: false
+
 # Calculate mean and standard deviation of price_difference
@@
 ```
-
-Figure 4: Distribution of Price Differences (Filtered).
-This histogram illustrates the frequency of price changes across products, with extreme values beyond three standard deviations removed for clarity.
-The concentration around zero suggests that small price adjustments are most common, while the presence of both negative and positive values indicates instances of both price increases and decreases.
-This distribution provides insight into the stability and variability of pricing across vendors.
-## Model Evaluation
-
-To evaluate our model’s performance, we assess the predictive accuracy of the Bayesian linear regression model on the test data.
-The model’s Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) provide insights into the model’s ability to predict `price_current` based on `price_old`, `vendor_name`, and `price_difference`.
-
-```{r}
-
-#### Read Data ####
-# Load only a sample of the data to reduce memory usage (e.g., 10% of the dataset)
-set.seed(853)
-
-analysis_data <- fread("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
-analysis_data <- analysis_data[sample(.N, size = 0.1 * .N)]
-
-# Split data into training and testing sets
-train_indices <- sample(seq_len(nrow(analysis_data)), size = 0.8 * nrow(analysis_data))
-analysis_data_train <- analysis_data[train_indices]
-analysis_data_test <- analysis_data[-train_indices]
-
-#### Model Data ####
-# Define a Bayesian regression model to predict `price_current` with fewer iterations and lighter priors
-
-model <- stan_glm(
-  formula = price_current ~ price_old + as.factor(vendor_name) + price_difference,
-  data = analysis_data_train,
-  family = gaussian(),
-  prior = normal(location = 0, scale = 5, autoscale = TRUE),      # Slightly broader prior
-  prior_intercept = normal(location = 0, scale = 5, autoscale = TRUE),
-  prior_aux = exponential(rate = 0.5, autoscale = TRUE),          # Less informative prior
-  iter = 500,                                                     # Reduce iterations
-  chains = 2,                                                     # Fewer chains for faster computation
-  seed = 853
-)
-
-#### Predict in Chunks ####
-# Define chunk size based on available memory
-chunk_size <- 20000  # Adjusted smaller chunk size for efficiency
-num_chunks <- ceiling(nrow(analysis_data_test) / chunk_size)
-
-# Initialize a vector to store predictions
-predictions <- numeric(nrow(analysis_data_test))
-
-# Loop through each chunk to generate predictions
-for (i in seq_len(num_chunks)) {
-  start <- ((i - 1) * chunk_size) + 1
-  end <- min(i * chunk_size, nrow(analysis_data_test))
-
-  # Generate predictions for the current chunk
-  predictions[start:end] <- predict(model, newdata = analysis_data_test[start:end, ])
-}
-
-#### Model Evaluation Metrics ####
-# Calculate Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE)
-mae <- mean(abs(predictions - analysis_data_test$price_current))
-rmse <- sqrt(mean((predictions - analysis_data_test$price_current)^2))
-
-# Store results in a data frame
-evaluation_metrics <- data.table(
-  Metric = c("Mean Absolute Error (MAE)", "Root Mean Squared Error (RMSE)"),
-  Value = c(round(mae, 2), round(rmse, 2))
-)
-
-# Display the results table
-print(evaluation_metrics)
-
-```
-
-Table 1: Model Evaluation Metrics.
-The table displays the MAE and RMSE for the model, reflecting the average and squared prediction errors, respectively.
-Lower values indicate better predictive accuracy, suggesting that the model effectively captures the relationship between previous prices, vendor identities, and price differences.
-
-These results demonstrate that vendor identity and previous prices are significant predictors of current prices in the Canadian grocery sector, with implications for understanding competitive strategies and consumer price sensitivity in the market.
 

### 2024-11-14 paper/paper.tex
Added fig-label

@@
 distribution provides insight into the stability and variability of
-pricing across vendors. \#\# Model Evaluation
-
-To evaluate our model's performance, we assess the predictive accuracy
-of the Bayesian linear regression model on the test data. The model's
-Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) provide
-insights into the model's ability to predict \texttt{price\_current}
-based on \texttt{price\_old}, \texttt{vendor\_name}, and
-\texttt{price\_difference}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\#\#\# Read Data \#\#\#\#}
-\CommentTok{\# Load only a sample of the data to reduce memory usage (e.g., 10\% of the dataset)}
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{853}\NormalTok{)}
-
-\NormalTok{analysis\_data }\OtherTok{\textless{}{-}} \FunctionTok{fread}\NormalTok{(}\StringTok{"C:/Users/eliza/Downloads/hammer{-}3{-}compressed/cleaned\_price\_changes.csv"}\NormalTok{)}
-\NormalTok{analysis\_data }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data[}\FunctionTok{sample}\NormalTok{(.N, }\AttributeTok{size =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ .N)]}
-
-\CommentTok{\# Split data into training and testing sets}
-\NormalTok{train\_indices }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\FunctionTok{seq\_len}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data)), }\AttributeTok{size =} \FloatTok{0.8} \SpecialCharTok{*} \FunctionTok{nrow}\NormalTok{(analysis\_data))}
-\NormalTok{analysis\_data\_train }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data[train\_indices]}
-\NormalTok{analysis\_data\_test }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data[}\SpecialCharTok{{-}}\NormalTok{train\_indices]}
-
-\DocumentationTok{\#\#\#\# Model Data \#\#\#\#}
-\CommentTok{\# Define a Bayesian regression model to predict \textasciigrave{}price\_current\textasciigrave{} with fewer iterations and lighter priors}
-\FunctionTok{library}\NormalTok{(rstanarm)}
-\NormalTok{model }\OtherTok{\textless{}{-}} \FunctionTok{stan\_glm}\NormalTok{(}
-  \AttributeTok{formula =}\NormalTok{ price\_current }\SpecialCharTok{\textasciitilde{}}\NormalTok{ price\_old }\SpecialCharTok{+} \FunctionTok{as.factor}\NormalTok{(vendor\_name) }\SpecialCharTok{+}\NormalTok{ price\_difference,}
-  \AttributeTok{data =}\NormalTok{ analysis\_data\_train,}
-  \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
-  \AttributeTok{prior =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \DecValTok{5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),      }\CommentTok{\# Slightly broader prior}
-  \AttributeTok{prior\_intercept =} \FunctionTok{normal}\NormalTok{(}\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \DecValTok{5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),}
-  \AttributeTok{prior\_aux =} \FunctionTok{exponential}\NormalTok{(}\AttributeTok{rate =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{autoscale =} \ConstantTok{TRUE}\NormalTok{),          }\CommentTok{\# Less informative prior}
-  \AttributeTok{iter =} \DecValTok{500}\NormalTok{,                                                     }\CommentTok{\# Reduce iterations}
-  \AttributeTok{chains =} \DecValTok{2}\NormalTok{,                                                     }\CommentTok{\# Fewer chains for faster computation}
-  \AttributeTok{seed =} \DecValTok{853}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-
-SAMPLING FOR MODEL 'continuous' NOW (CHAIN 1).
-Chain 1: 
-Chain 1: Gradient evaluation took 0.000105 seconds
-Chain 1: 1000 transitions using 10 leapfrog steps per transition would take 1.05 seconds.
-Chain 1: Adjust your expectations accordingly!
-Chain 1: 
-Chain 1: 
-Chain 1: Iteration:   1 / 500 [  0%]  (Warmup)
-Chain 1: Iteration:  50 / 500 [ 10%]  (Warmup)
-Chain 1: Iteration: 100 / 500 [ 20%]  (Warmup)
-Chain 1: Iteration: 150 / 500 [ 30%]  (Warmup)
-Chain 1: Iteration: 200 / 500 [ 40%]  (Warmup)
-Chain 1: Iteration: 250 / 500 [ 50%]  (Warmup)
-Chain 1: Iteration: 251 / 500 [ 50%]  (Sampling)
-Chain 1: Iteration: 300 / 500 [ 60%]  (Sampling)
-Chain 1: Iteration: 350 / 500 [ 70%]  (Sampling)
-Chain 1: Iteration: 400 / 500 [ 80%]  (Sampling)
-Chain 1: Iteration: 450 / 500 [ 90%]  (Sampling)
-Chain 1: Iteration: 500 / 500 [100%]  (Sampling)
-Chain 1: 
-Chain 1:  Elapsed Time: 0.056 seconds (Warm-up)
-Chain 1:                3.902 seconds (Sampling)
-Chain 1:                3.958 seconds (Total)
-Chain 1:
-
-SAMPLING FOR MODEL 'continuous' NOW (CHAIN 2).
-Chain 2: 
-Chain 2: Gradient evaluation took 2.5e-05 seconds
-Chain 2: 1000 transitions using 10 leapfrog steps per transition would take 0.25 seconds.
-Chain 2: Adjust your expectations accordingly!
-Chain 2: 
-Chain 2: 
-Chain 2: Iteration:   1 / 500 [  0%]  (Warmup)
-Chain 2: Iteration:  50 / 500 [ 10%]  (Warmup)
-Chain 2: Iteration: 100 / 500 [ 20%]  (Warmup)
-Chain 2: Iteration: 150 / 500 [ 30%]  (Warmup)
-Chain 2: Iteration: 200 / 500 [ 40%]  (Warmup)
-Chain 2: Iteration: 250 / 500 [ 50%]  (Warmup)
-Chain 2: Iteration: 251 / 500 [ 50%]  (Sampling)
-Chain 2: Iteration: 300 / 500 [ 60%]  (Sampling)
-Chain 2: Iteration: 350 / 500 [ 70%]  (Sampling)
-Chain 2: Iteration: 400 / 500 [ 80%]  (Sampling)
-Chain 2: Iteration: 450 / 500 [ 90%]  (Sampling)
-Chain 2: Iteration: 500 / 500 [100%]  (Sampling)
-Chain 2: 
-Chain 2:  Elapsed Time: 0.036 seconds (Warm-up)
-Chain 2:                3.936 seconds (Sampling)
-Chain 2:                3.972 seconds (Total)
-Chain 2: 
-\end{verbatim}
-
-\begin{verbatim}
-Warning: There were 1 chains where the estimated Bayesian Fraction of Missing Information was low. See
-https://mc-stan.org/misc/warnings.html#bfmi-low
-\end{verbatim}
-
-\begin{verbatim}
-Warning: Examine the pairs() plot to diagnose sampling problems
-\end{verbatim}
-
-\begin{verbatim}
-Warning: The largest R-hat is 1.6, indicating chains have not mixed.
-Running the chains for more iterations may help. See
-https://mc-stan.org/misc/warnings.html#r-hat
-\end{verbatim}
-
-\begin{verbatim}
-Warning: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.
-Running the chains for more iterations may help. See
-https://mc-stan.org/misc/warnings.html#bulk-ess
-\end{verbatim}
-
-\begin{verbatim}
-Warning: Tail Effective Samples Size (ESS) is too low, indicating posterior variances and tail quantiles may be unreliable.
-Running the chains for more iterations may help. See
-https://mc-stan.org/misc/warnings.html#tail-ess
-\end{verbatim}
-
-\begin{verbatim}
-Warning: Markov chains did not converge! Do not analyze results!
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\#\#\# Predict in Chunks \#\#\#\#}
-\CommentTok{\# Define chunk size based on available memory}
-\NormalTok{chunk\_size }\OtherTok{\textless{}{-}} \DecValTok{20000}  \CommentTok{\# Adjusted smaller chunk size for efficiency}
-\NormalTok{num\_chunks }\OtherTok{\textless{}{-}} \FunctionTok{ceiling}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data\_test) }\SpecialCharTok{/}\NormalTok{ chunk\_size)}
-
-\CommentTok{\# Initialize a vector to store predictions}
-\NormalTok{predictions }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\FunctionTok{nrow}\NormalTok{(analysis\_data\_test))}
-
-\CommentTok{\# Loop through each chunk to generate predictions}
-\ControlFlowTok{for}\NormalTok{ (i }\ControlFlowTok{in} \FunctionTok{seq\_len}\NormalTok{(num\_chunks)) \{}
-\NormalTok{  start }\OtherTok{\textless{}{-}}\NormalTok{ ((i }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ chunk\_size) }\SpecialCharTok{+} \DecValTok{1}
-\NormalTok{  end }\OtherTok{\textless{}{-}} \FunctionTok{min}\NormalTok{(i }\SpecialCharTok{*}\NormalTok{ chunk\_size, }\FunctionTok{nrow}\NormalTok{(analysis\_data\_test))}
-
-  \CommentTok{\# Generate predictions for the current chunk}
-\NormalTok{  predictions[start}\SpecialCharTok{:}\NormalTok{end] }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(model, }\AttributeTok{newdata =}\NormalTok{ analysis\_data\_test[start}\SpecialCharTok{:}\NormalTok{end, ])}
-\NormalTok{\}}
-
-\DocumentationTok{\#\#\#\# Model Evaluation Metrics \#\#\#\#}
-\CommentTok{\# Calculate Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE)}
-\NormalTok{mae }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{abs}\NormalTok{(predictions }\SpecialCharTok{{-}}\NormalTok{ analysis\_data\_test}\SpecialCharTok{$}\NormalTok{price\_current))}
-\NormalTok{rmse }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(}\FunctionTok{mean}\NormalTok{((predictions }\SpecialCharTok{{-}}\NormalTok{ analysis\_data\_test}\SpecialCharTok{$}\NormalTok{price\_current)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{))}
-
-\CommentTok{\# Store results in a data frame}
-\NormalTok{evaluation\_metrics }\OtherTok{\textless{}{-}} \FunctionTok{data.table}\NormalTok{(}
-  \AttributeTok{Metric =} \FunctionTok{c}\NormalTok{(}\StringTok{"Mean Absolute Error (MAE)"}\NormalTok{, }\StringTok{"Root Mean Squared Error (RMSE)"}\NormalTok{),}
-  \AttributeTok{Value =} \FunctionTok{c}\NormalTok{(}\FunctionTok{round}\NormalTok{(mae, }\DecValTok{2}\NormalTok{), }\FunctionTok{round}\NormalTok{(rmse, }\DecValTok{2}\NormalTok{))}
-\NormalTok{)}
-
-\CommentTok{\# Display the results table}
-\FunctionTok{print}\NormalTok{(evaluation\_metrics)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-                           Metric Value
-                           <char> <num>
-1:      Mean Absolute Error (MAE)  6.74
-2: Root Mean Squared Error (RMSE) 17.98
-\end{verbatim}
-
-Table 1: Model Evaluation Metrics. The table displays the MAE and RMSE
-for the model, reflecting the average and squared prediction errors,
-respectively. Lower values indicate better predictive accuracy,
-suggesting that the model effectively captures the relationship between
-previous prices, vendor identities, and price differences.
-
-These results demonstrate that vendor identity and previous prices are
-significant predictors of current prices in the Canadian grocery sector,
-with implications for understanding competitive strategies and consumer
-price sensitivity in the market.
+pricing across vendors.
 

### 2024-11-14 paper/paper.qmd
Updated intro, measurement, and abstract

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Analyzing changes in Canadian grocery prices"
+# subtitle: "My subtitle if needed"
 author: 
-  - First author
-  - Another author
+  - Abdullah Motasim
+  - Elizabeth Luong
 thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper investigates grocery pricing trends across Canada using SQL analysis on the 'Project Hammer' dataset, which contains detailed vendor-specific price data.
+Focusing on statistical relationships and potential pricing biases, we explore factors driving price variations across different regions and vendors.
+A Bayesian regression model is applied to understand correlations, accounting for limitations and highlighting areas where causation cannot be assumed.
+Findings reveal competitive dynamics within Canadian grocery markets, implications for consumers, and recommendations for future research into market and pricing behavior across vendors."
 format: pdf
@@
 
-Overview paragraph
+The rising cost of groceries in Canada has drawn significant public attention, impacting consumer budgets and prompting regulatory interest.
+While inflation and supply disruptions are often cited as contributors, the specifics of how major Canadian grocery vendors set and adjust prices remain underexplored.
+This study addresses this gap by analyzing data from Project Hammer, a dataset that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
+Through examining these data, we aim to uncover factors driving grocery price dynamics and assess the degree to which vendor-specific pricing may reflect competitive or non-competitive practices.
 
-Estimand paragraph
+Using a Bayesian regression model, our analysis identifies prior prices, vendor identity, and recent price changes as significant determinants of current grocery prices.
+The consistency in historical pricing indicates relative stability across product categories, while unique vendor patterns suggest differing market strategies, likely shaped by operational costs and customer demographics.
 
-Results paragraph
-
-Why it matters paragraph
-
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+Our findings have practical implications for both consumers and policymakers: consumers can use vendor-specific insights to make informed purchasing decisions, while policymakers can leverage this analysis to support regulatory strategies that promote competition and price transparency.
+The paper is structured as follows: Section 2 reviews the dataset, Section 3 describes the Bayesian model, Section 4 presents the results, and Section 5 discusses broader implications and future research directions.
 
@@
 
-## Outcome variables
+Each entry in the dataset represents a captured moment in time where prices, products, and vendor information were recorded.
+The measurement process involved regular data collection from the online pricing systems of the following prominent grocery chains: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
+Each entry captures variables such as:  product name, vendor name, and both current and previous prices.
+The Project Hammer initiative leveraged web scraping technology to ensure a consistent and standardized entry format for all retailers, documenting prices at regular intervals.
+This approach ensures that price points reflect real-time data rather than historical estimates or annual averages, providing a close approximation of consumer experiences.
+The data set records the price of each grocery item in Canadian dollars, including items from broad categories such as fresh produce, dairy, pantry goods, and household essentials, to present a holistic view of grocery costs.
 
-## Predictor variables
+<!-- Each row entry in Project Hammer represents a unique product listing per vendor, featuring five main columns: product_name, vendor_name, price_current, price_old, and price_difference.
+The price_difference column is particularly useful for observing temporal price changes, while the price_old variable provides context for price continuity or volatility.
+The structured approach used in capturing this data allows us to transform general observations of grocery pricing trends into quantifiable entries that can be statistically analyzed to reveal underlying patterns in pricing strategies across different vendors and products.
+-->
+
+Essentially, the real world phenomenon we observed was the website of each vendor with a listing of their products for that week, we turned this phenomenon into an entry within out data set with the use of a screen-scrape of the website UI.
+This means a HTTP request to load the page was sent to the desired vendors website on a specific day, than the returned HTML was parsed to extract specific content such as text, images, prices, etc.
+These captured features were than utilized to fill out the corresponding columns within the data set.
 
@@
 ```{r}
-#| label: fig-avg-curr-price
-#| fig-cap: Average Current Price by Vendor.
-This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD).
-Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.
+#| label: fig1
+#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
 #| echo: false
@@
 # Load the cleaned dataset
-analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv", show_col_types = FALSE)
-
-# Generate plot for Average Current Price by Vendor
-vendor_price_plot <- analysis_data %>%
-  group_by(vendor_name) %>%
-  summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
-  ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
-  geom_bar(stat = "identity", fill = "skyblue") +
-  coord_flip() +
-  labs(
-    title = "Average Current Price by Vendor",
-    x = "Vendor",
-    y = "Average Current Price (CAD)"
-  ) +
-  theme_minimal()
-
-# Display the plot
-vendor_price_plot
+# analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv", show_col_types = FALSE)
+# 
+# # Generate plot for Average Current Price by Vendor
+# vendor_price_plot <- analysis_data %>%
+#   group_by(vendor_name) %>%
+#   summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
+#   ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
+#   geom_bar(stat = "identity", fill = "skyblue") +
+#   coord_flip() +
+#   labs(
+#     title = "Average Current Price by Vendor",
+#     x = "Vendor",
+#     y = "Average Current Price (CAD)"
+#   ) +
+#   theme_minimal()
+# 
+# # Display the plot
+# vendor_price_plot
 
 ```
+
+Figure 1: Average Current Price by Vendor.
+This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD).
+Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.
 
@@
 ```{r}
-#| label: fig-avg-price-diff
-#| fig-cap: Average Price Difference by Vendor.
-This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments.
-Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.
+#| label: fig2
+#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
 #| echo: false
@@
 # Average price difference by vendor
-price_diff_plot <- analysis_data %>%
-  group_by(vendor_name) %>%
-  summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
-  ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
-  geom_bar(stat = "identity", fill = "coral") +
-  coord_flip() +
-  labs(
-    title = "Average Price Difference by Vendor",
-    x = "Vendor",
-    y = "Average Price Difference (CAD)"
-  ) +
-  theme_minimal()
-price_diff_plot
+# price_diff_plot <- analysis_data %>%
+#   group_by(vendor_name) %>%
+#   summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
+#   ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
+#   geom_bar(stat = "identity", fill = "coral") +
+#   coord_flip() +
+#   labs(
+#     title = "Average Price Difference by Vendor",
+#     x = "Vendor",
+#     y = "Average Price Difference (CAD)"
+#   ) +
+#   theme_minimal()
+# price_diff_plot
 
 ```
+
+Figure 2: Average Price Difference by Vendor.
+This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments.
+Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.
 
@@
 ```{r}
-#| label: fig-past-curr
-#| fig-cap: Relationship Between Previous and Current Prices.
-This scatter plot with a linear regression line shows the correlation between previous and current prices.
-A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
+#| label: fig3
+#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
 #| echo: false
@@
 # Scatter plot of previous price vs. current price
-price_relationship_plot <- ggplot(analysis_data, aes(x = price_old, y = price_current)) +
-  geom_point(alpha = 0.4, color = "blue") +
-  geom_smooth(method = "lm", color = "red") +
-  labs(
-    title = "Relationship Between Previous and Current Prices",
-    x = "Previous Price (CAD)",
-    y = "Current Price (CAD)"
-  ) +
-  theme_minimal()
-price_relationship_plot
+# price_relationship_plot <- ggplot(analysis_data, aes(x = price_old, y = price_current)) +
+#   geom_point(alpha = 0.4, color = "blue") +
+#   geom_smooth(method = "lm", color = "red") +
+#   labs(
+#     title = "Relationship Between Previous and Current Prices",
+#     x = "Previous Price (CAD)",
+#     y = "Current Price (CAD)"
+#   ) +
+#   theme_minimal()
+# price_relationship_plot
 
 ```
+Figure 3: Relationship Between Previous and Current Prices.
+This scatter plot with a linear regression line shows the correlation between previous and current prices.
+A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
 
@@
 ```{r}
-#| label: fig-dist-price
-#| fig-cap: Relationship Between Previous and Current Prices.
-This scatter plot with a linear regression line shows the correlation between previous and current prices.
-A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
+#| label: fig4
+#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
 #| echo: false
@@
 # Calculate mean and standard deviation of price_difference
-mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
-sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)
-
-# Filter out extreme values beyond 3 standard deviations
-filtered_data <- analysis_data %>%
-  filter(price_difference >= (mean_diff - 3 * sd_diff),
-         price_difference <= (mean_diff + 3 * sd_diff))
-
-# Plot histogram with filtered data
-price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
-  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
-  labs(
-    title = "Distribution of Price Differences (Filtered)",
-    x = "Price Difference (CAD)",
-    y = "Frequency"
-  ) +
-  theme_minimal()
-price_difference_hist
+# mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
+# sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)
+# 
+# # Filter out extreme values beyond 3 standard deviations
+# filtered_data <- analysis_data %>%
+#   filter(price_difference >= (mean_diff - 3 * sd_diff),
+#          price_difference <= (mean_diff + 3 * sd_diff))
+# 
+# # Plot histogram with filtered data
+# price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
+#   geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
+#   labs(
+#     title = "Distribution of Price Differences (Filtered)",
+#     x = "Price Difference (CAD)",
+#     y = "Frequency"
+#   ) +
+#   theme_minimal()
+# price_difference_hist
 
 ```
+
+Figure 4: Distribution of Price Differences (Filtered).
+This histogram illustrates the frequency of price changes across products, with extreme values beyond three standard deviations removed for clarity.
+The concentration around zero suggests that small price adjustments are most common, while the presence of both negative and positive values indicates instances of both price increases and decreases.
+This distribution provides insight into the stability and variability of pricing across vendors.
+## Model Evaluation
+
+To evaluate our model’s performance, we assess the predictive accuracy of the Bayesian linear regression model on the test data.
+The model’s Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE) provide insights into the model’s ability to predict `price_current` based on `price_old`, `vendor_name`, and `price_difference`.
+
+```{r}
+#| label: fig5
+#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
+#| echo: false
+#| warning: false
+#| message: false
+
+#### Read Data ####
+# Load only a sample of the data to reduce memory usage (e.g., 10% of the dataset)
+# set.seed(853)
+# 
+# analysis_data <- fread("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv")
+# analysis_data <- analysis_data[sample(.N, size = 0.1 * .N)]
+# 
+# # Split data into training and testing sets
+# train_indices <- sample(seq_len(nrow(analysis_data)), size = 0.8 * nrow(analysis_data))
+# analysis_data_train <- analysis_data[train_indices]
+# analysis_data_test <- analysis_data[-train_indices]
+# 
+# #### Model Data ####
+# # Define a Bayesian regression model to predict `price_current` with fewer iterations and lighter priors
+# 
+# model <- stan_glm(
+#   formula = price_current ~ price_old + as.factor(vendor_name) + price_difference,
+#   data = analysis_data_train,
+#   family = gaussian(),
+#   prior = normal(location = 0, scale = 5, autoscale = TRUE),      # Slightly broader prior
+#   prior_intercept = normal(location = 0, scale = 5, autoscale = TRUE),
+#   prior_aux = exponential(rate = 0.5, autoscale = TRUE),          # Less informative prior
+#   iter = 500,                                                     # Reduce iterations
+#   chains = 2,                                                     # Fewer chains for faster computation
+#   seed = 853
+# )
+# 
+# #### Predict in Chunks ####
+# # Define chunk size based on available memory
+# chunk_size <- 20000  # Adjusted smaller chunk size for efficiency
+# num_chunks <- ceiling(nrow(analysis_data_test) / chunk_size)
+# 
+# # Initialize a vector to store predictions
+# predictions <- numeric(nrow(analysis_data_test))
+# 
+# # Loop through each chunk to generate predictions
+# for (i in seq_len(num_chunks)) {
+#   start <- ((i - 1) * chunk_size) + 1
+#   end <- min(i * chunk_size, nrow(analysis_data_test))
+#   
+#   # Generate predictions for the current chunk
+#   predictions[start:end] <- predict(model, newdata = analysis_data_test[start:end, ])
+# }
+# 
+# #### Model Evaluation Metrics ####
+# # Calculate Mean Absolute Error (MAE) and Root Mean Squared Error (RMSE)
+# mae <- mean(abs(predictions - analysis_data_test$price_current))
+# rmse <- sqrt(mean((predictions - analysis_data_test$price_current)^2))
+# 
+# # Store results in a data frame
+# evaluation_metrics <- data.table(
+#   Metric = c("Mean Absolute Error (MAE)", "Root Mean Squared Error (RMSE)"),
+#   Value = c(round(mae, 2), round(rmse, 2))
+# )
+# 
+# # Display the results table
+# print(evaluation_metrics)
+
+```
+
+Table 1: Model Evaluation Metrics.
+The table displays the MAE and RMSE for the model, reflecting the average and squared prediction errors, respectively.
+Lower values indicate better predictive accuracy, suggesting that the model effectively captures the relationship between previous prices, vendor identities, and price differences.
+
+These results demonstrate that vendor identity and previous prices are significant predictors of current prices in the Canadian grocery sector, with implications for understanding competitive strategies and consumer price sensitivity in the market.
 
@@
 
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
 \newpage

### 2024-11-14 paper/paper.qmd

@@
 ```{r}
-#| label: fig1
-#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
+#| label: fig-avg-curr-price
+#| fig-cap: Average Current Price by Vendor.
+This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD).
+Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.
 #| echo: false
@@
 ```{r}
-#| label: fig2
-#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
+#| label: fig-avg-price-diff
+#| fig-cap: Average Price Difference by Vendor.
+This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments.
+Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.
 #| echo: false
@@
 ```{r}
-#| label: fig3
-#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
+#| label: fig-past-curr
+#| fig-cap: Relationship Between Previous and Current Prices.
+This scatter plot with a linear regression line shows the correlation between previous and current prices.
+A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
 #| echo: false
@@
 ```{r}
-#| label: fig4
-#| fig-cap: Each boxplot shows the distribution of numeric reliability grades for the top 5 pollsters, with individual poll grades represented by jittered points.
+#| label: fig-dist-price
+#| fig-cap: Relationship Between Previous and Current Prices.
+This scatter plot with a linear regression line shows the correlation between previous and current prices.
+A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
 #| echo: false
@@
 
-Figure 4: Distribution of Price Differences (Filtered).
+<!-- Figure 4: Distribution of Price Differences (Filtered).
 This histogram illustrates the frequency of price changes across products, with extreme values beyond three standard deviations removed for clarity.
@@
 This distribution provides insight into the stability and variability of pricing across vendors.
+-->
+
 ## Model Evaluation
@@
 
+=======
+>>>>>>> 3b05c34e7a7a3720d0e488dcd31eaaaa4c292cae
+
 # Discussion

### 2024-11-14 paper/paper.qmd
Added sub-sections

--- 2024-11-14 paper/paper.qmd
@@
 
-The results of our analysis of the Project Hammer dataset explain important findings about pricing trends across major Canadian grocery vendors.
+The results of our analysis of the Project Hammer dataset shows significant findings into pricing trends across major Canadian grocery vendors.
 By examining factors such as prior pricing (`price_old`), vendor identities (`vendor_name`), and recent price changes (`price_difference`), we assess how these variables affect current prices (`price_current`).
@@
 ```
+@fig-avg-curr-price displays the average current prices across vendors, revealing notable pricing differences.
+Metro and Galleria, for example, tend to have higher average prices, while No Frills and Walmart are on the lower end.
+These discrepancies suggest differing market positioning, with some vendors potentially catering to more price-sensitive customers.
+
+**Implications for Consumers**: Budget-conscious consumers might prefer Walmart or No Frills for lower average prices, while those seeking specific product selections or brands might choose Metro or Galleria despite higher prices.
 
 ## Price Difference Analysis
-
-We further analyze `price_difference`, which captures the change between the current price (`price_current`) and the previous price (`price_old`).
-This variable provides insights into recent price adjustments across vendors.
+We further analyze price_difference, which captures the change between the current price (`price_current`) and the previous price (`price_old`).
+This variable provides findings into recent price adjustments across vendors.
 
@@
 ```
+@fig-avg-price-diff illustrates the average price differences by vendor.
+Retailers like Walmart and Galleria exhibit negative average price differences, suggesting a tendency to reduce prices, likely through competitive discounts.
+On the other hand, Loblaw shows relatively stable or small positive differences, suggesting less frequent or minor price adjustments.
+
+**Competitive Strategy**: Vendors with negative average differences may be aiming to attract customers through regular discounts, while vendors with stable or positive differences might rely on consistent pricing for brand loyalty.
 
 ## Relationship Between Previous and Current Prices
-
-We assess the relationship between previous prices (`price_old`) and current prices (`price_current`) to understand price consistency and potential price elasticity across products.
+To understand price consistency and elasticity, we assess the relationship between previous prices (`price_old`) and current prices (`price_current`).
 
@@
 ```
+@fig-past-curr shows a strong positive relationship between previous prices and current prices.
+This pattern indicates that higher past prices are generally associated with higher current prices, implying stability in grocery pricing over time.
+
+**Price Elasticity Findings**: The linear trend suggests low price elasticity, where previous prices predict current prices with relative stability.
+This may reflect supply chain consistency and retailers’ cautious approach to pricing changes.
 
 ## Distribution of Price Differences
-
 To further explore pricing dynamics, we examine the distribution of `price_difference`, capturing both price increases and decreases across the dataset.
@@
 ```
+@fig-dist-price (filtered to remove extreme values), shows that most price changes are small, with values clustered around zero.
+Both positive and negative price differences are present, indicating a balance between price increases and decreases across vendors.
+
+**Market Stability**: The concentration around zero suggests a generally stable market, with moderate pricing adjustments.
+While some vendors increase prices, others offer discounts, contributing to a balanced market.
 
@@
 
-## Interpretation of Grocery Pricing Dynamics
-This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset.
-Our Bayesian regression model reveals that prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`) play significant roles in determining current prices.
-The influence of previous prices indicates price continuity across time, suggesting that grocery items tend to exhibit relatively stable pricing patterns with minor adjustments.
-Vendor-specific effects highlight the role of retailer pricing strategies, with some vendors maintaining consistently higher or lower prices.
-These vendor variations may reflect differences in market positioning, operational costs, or competitive strategies, impacting price consistency across the grocery sector.
+## Correlation vs. Causation
+This analysis utilizes Bayesian regression model to identify factors correlated with current grocery prices, including prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`).
+While the model shows that historical prices and vendor-specific effects are strongly associated with current pricing, it’s essential to interpret these relationships cautiously in terms of causation.
 
-Additionally, the presence of both positive and negative values in the `price_difference` variable illustrates the prevalence of both price increases and decreases, underscoring the dynamic nature of grocery pricing in response to external market forces.
-External factors, such as supply chain disruptions, inflation, and seasonal demand fluctuations, likely contribute to the observed variability.
-This model highlights the complexity of pricing dynamics and the importance of historical pricing and vendor-specific factors in shaping current grocery prices across major Canadian retailers.
+The correlation between previous and current prices likely reflects price continuity rather than a causal mechanism.
+For example, stable pricing over time may arise from vendor policies or industry norms rather than from any inherent property of the product or market structure.
+Similarly, vendor identity correlates with price differences, but this does not imply that being a particular vendor causes higher or lower prices.
+Instead, vendor-specific prices likely reflect underlying operational costs, competitive positioning, or branding strategies.
+Caution is warranted in inferring causation, as unobserved factors—such as cost structures or consumer loyalty—may influence both vendor identity and pricing patterns.
 
-## Implications for Competition and Consumer Costs
+The presence of both price increases and decreases (reflected in `price_difference`) further highlights the dynamic nature of grocery pricing.
+While these changes are correlated with the final prices observed, it would be speculative to conclude that specific price fluctuations directly cause current price levels.
+External factors like inflation, seasonal demand, and supply chain disruptions may drive these observed correlations, underscoring the need for a cautious interpretation of causality.
 
-The findings from Project Hammer have broader implications for both competition in the grocery sector and consumer expenses.
-Vendor-specific price differences suggest that some retailers adopt competitive pricing strategies to attract cost-sensitive consumers, while others may leverage brand loyalty or perceived quality to justify higher prices.
-In this context, price-sensitive consumers may benefit from exploring pricing variations across vendors to find the best deals, while retailers may face pressure to adjust their prices to remain competitive.
+## Missing Data
 
-The stability of historical prices as a predictor of current prices also highlights potential concerns about price rigidity in certain product categories.
-This rigidity may limit the effectiveness of competition, as vendors might rely on historical price baselines rather than actively adjusting prices in response to market demand.
-Policymakers interested in promoting competition in the grocery sector could consider encouraging more frequent pricing updates or transparency in pricing strategies, particularly for staple items that heavily impact household budgets.
+The Project Hammer dataset contains some limitations in data coverage, which may impact the completeness and accuracy of our findings.
+Specifically, the data does not include smaller or regional grocery stores, which may have distinct pricing strategies and could offer lower or more competitive prices in certain areas.
+The absence of these smaller players may lead to an incomplete picture of the overall grocery market, as price variations across different types of stores are not captured.
 
-## Limitations of Data and Model
+Additionally, missing data on certain product categories or regions within vendors limits the granularity of the analysis.
+For example, without category-specific data (e.g., dairy, produce, or packaged goods), it’s challenging to examine whether price dynamics vary significantly across product types.
+This lack of detail may obscure category-level trends, which could be particularly relevant for staple items that heavily influence household spending.
+Future data collection efforts could focus on capturing a wider range of store types and product details to provide a more representative dataset.
 
-This analysis has several limitations that may impact the accuracy and comprehensiveness of our findings.
-First, the dataset is limited to major grocery chains, which may not capture price variations across smaller or regional stores that could offer different pricing structures.
-Additionally, the lack of granular product-level data limits our ability to analyze category-specific trends (e.g., dairy versus produce) or regional differences within the same vendor.
+Finally, missing information on factors like seasonal influences, promotions, or regional economic conditions means that some relevant variables are absent from the model.
+These factors could help explain variations in grocery pricing, particularly for products sensitive to external factors (e.g., fresh produce affected by seasonality).
+Addressing these data gaps could significantly improve the model’s ability to capture the full range of influences on grocery prices.
 
-The model’s focus on historical prices, vendor identity, and price differences also excludes potential external factors that may influence pricing, such as seasonal demand, promotions, or supply chain constraints.
-Including additional variables, such as the frequency of price adjustments or regional economic indicators, could enhance the model’s predictive power and provide a more comprehensive view of grocery pricing dynamics.
+## Sources of Bias
 
-## Future Research Direction
+The Project Hammer dataset and the corresponding analysis may contain several sources of bias that could affect the interpretation of results.
+One potential source of bias stems from the focus on major grocery chains, which may not accurately represent pricing dynamics across smaller, independent stores.
+Since major vendors typically have broader operational reach and standardized pricing strategies, the findings may be biased toward the practices of these larger corporations, potentially overestimating price stability or ignoring competitive pricing seen in smaller markets.
 
-Future research could build on this work by incorporating a wider range of variables, including product-specific attributes, seasonal factors, and promotional data, to capture a broader spectrum of influences on grocery prices.
-Exploring how demographic factors, such as income levels in different regions, correlate with price patterns across vendors could offer insights into the socio-economic impacts of grocery pricing on different communities.
+Another possible bias comes from the data’s retrospective nature, focusing on historical pricing as a predictor of current prices.
+This approach may inherently favor price continuity and overlook the role of short-term market disruptions, promotions, or external shocks.
+For instance, a focus on historical prices could understate the impact of sudden events (like supply chain breakdowns or economic crises) that may cause sharp price changes.
+This form of selection bias could lead to an overly conservative estimate of price volatility in the model.
 
-Increasing the model’s granularity to analyze price dynamics at the regional or product-category level would provide a deeper understanding of how specific items are priced within the same store or across regions.
-Additionally, expanding the analysis to include other sectors, such as online grocery prices or specialty stores, could provide a more holistic view of the Canadian grocery market.
-
-Incorporating machine learning techniques could also improve the model's adaptability to rapidly changing market conditions, allowing for more nuanced forecasts of pricing trends.
-This approach could be particularly useful in the face of ongoing challenges such as inflation and supply chain disruptions, offering a valuable tool for policymakers and consumers interested in understanding and mitigating grocery costs in Canada.
+Furthermore, vendor-specific effects could introduce bias if certain vendors follow distinct pricing philosophies, such as premium or discount positioning.
+If these vendors have unique market positions that attract particular types of consumers, the model may reflect not only price dynamics but also underlying consumer segmentation.
+This brand-driven bias could distort interpretations, particularly if certain vendors consistently maintain higher or lower prices due to factors outside competitive dynamics (e.g., brand loyalty or perceived product quality).
 

### 2024-11-14 paper/paper.qmd
Updated paper

--- 2024-11-14 paper/paper.qmd
@@
 While inflation and supply disruptions are often cited as contributors, the specifics of how major Canadian grocery vendors set and adjust prices remain underexplored.
-This study addresses this gap by analyzing data from Project Hammer, a dataset that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
+This study addresses this gap by analyzing data from Project Hammer, a data set that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
 Through examining these data, we aim to uncover factors driving grocery price dynamics and assess the degree to which vendor-specific pricing may reflect competitive or non-competitive practices.
@@
 The consistency in historical pricing indicates relative stability across product categories, while unique vendor patterns suggest differing market strategies, likely shaped by operational costs and customer demographics.
+Our findings have practical implications for both consumers and policymakers: consumers can use vendor-specific insights to make informed purchasing decisions, while policymakers can leverage this analysis to support regulatory strategies that promote competition and price transparency.
 
-Our findings have practical implications for both consumers and policymakers: consumers can use vendor-specific insights to make informed purchasing decisions, while policymakers can leverage this analysis to support regulatory strategies that promote competition and price transparency.
-The paper is structured as follows: Section 2 reviews the dataset, Section 3 describes the Bayesian model, Section 4 presents the results, and Section 5 discusses broader implications and future research directions.
+The paper is structured as follows: @sec-data discusses the data types included in the raw data, the cleaning process for the data, and the reason for selecting the data set we did.
+@sec-model discusses model specification and justification for utilizing a Bayesian linear regression model.
+@sec-results presents the trends and correlations between different variables utilizing tabular and graphical means.
+@sec-discussion discusses the results of @sec-results going into detail on what the simulation results can tell us about grocery prices in Canada as well as discussing missing data and sources of bias.
 
@@
 
-Each entry in the dataset represents a captured moment in time where prices, products, and vendor information were recorded.
+Each entry in the data set represents a captured moment in time where prices, products, and vendor information were recorded.
 The measurement process involved regular data collection from the online pricing systems of the following prominent grocery chains: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
@@
 
-# Model
+# Model {#sec-model}
 
@@
 
-# Results
+# Results {#sec-results}
 
-The results of our analysis of the Project Hammer dataset shows significant findings into pricing trends across major Canadian grocery vendors.
+The results of our analysis of the Project Hammer dataset explain important findings about pricing trends across major Canadian grocery vendors.
 By examining factors such as prior pricing (`price_old`), vendor identities (`vendor_name`), and recent price changes (`price_difference`), we assess how these variables affect current prices (`price_current`).
@@
 ```
-@fig-avg-curr-price displays the average current prices across vendors, revealing notable pricing differences.
-Metro and Galleria, for example, tend to have higher average prices, while No Frills and Walmart are on the lower end.
-These discrepancies suggest differing market positioning, with some vendors potentially catering to more price-sensitive customers.
-
-**Implications for Consumers**: Budget-conscious consumers might prefer Walmart or No Frills for lower average prices, while those seeking specific product selections or brands might choose Metro or Galleria despite higher prices.
 
 ## Price Difference Analysis
-We further analyze price_difference, which captures the change between the current price (`price_current`) and the previous price (`price_old`).
-This variable provides findings into recent price adjustments across vendors.
+
+We further analyze `price_difference`, which captures the change between the current price (`price_current`) and the previous price (`price_old`).
+This variable provides insights into recent price adjustments across vendors.
 
@@
 ```
-@fig-avg-price-diff illustrates the average price differences by vendor.
-Retailers like Walmart and Galleria exhibit negative average price differences, suggesting a tendency to reduce prices, likely through competitive discounts.
-On the other hand, Loblaw shows relatively stable or small positive differences, suggesting less frequent or minor price adjustments.
-
-**Competitive Strategy**: Vendors with negative average differences may be aiming to attract customers through regular discounts, while vendors with stable or positive differences might rely on consistent pricing for brand loyalty.
 
 ## Relationship Between Previous and Current Prices
-To understand price consistency and elasticity, we assess the relationship between previous prices (`price_old`) and current prices (`price_current`).
+
+We assess the relationship between previous prices (`price_old`) and current prices (`price_current`) to understand price consistency and potential price elasticity across products.
 
@@
 ```
-@fig-past-curr shows a strong positive relationship between previous prices and current prices.
-This pattern indicates that higher past prices are generally associated with higher current prices, implying stability in grocery pricing over time.
-
-**Price Elasticity Findings**: The linear trend suggests low price elasticity, where previous prices predict current prices with relative stability.
-This may reflect supply chain consistency and retailers’ cautious approach to pricing changes.
 
 ## Distribution of Price Differences
+
 To further explore pricing dynamics, we examine the distribution of `price_difference`, capturing both price increases and decreases across the dataset.
@@
 ```
-@fig-dist-price (filtered to remove extreme values), shows that most price changes are small, with values clustered around zero.
-Both positive and negative price differences are present, indicating a balance between price increases and decreases across vendors.
-
-**Market Stability**: The concentration around zero suggests a generally stable market, with moderate pricing adjustments.
-While some vendors increase prices, others offer discounts, contributing to a balanced market.
 
@@
 
-=======
->>>>>>> 3b05c34e7a7a3720d0e488dcd31eaaaa4c292cae
+# Discussion {sec-discussion}
 
-# Discussion
+## Interpretation of Grocery Pricing Dynamics
+This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset.
+Our Bayesian regression model reveals that prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`) play significant roles in determining current prices.
+The influence of previous prices indicates price continuity across time, suggesting that grocery items tend to exhibit relatively stable pricing patterns with minor adjustments.
+Vendor-specific effects highlight the role of retailer pricing strategies, with some vendors maintaining consistently higher or lower prices.
+These vendor variations may reflect differences in market positioning, operational costs, or competitive strategies, impacting price consistency across the grocery sector.
 
-## Correlation vs. Causation
-This analysis utilizes Bayesian regression model to identify factors correlated with current grocery prices, including prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`).
-While the model shows that historical prices and vendor-specific effects are strongly associated with current pricing, it’s essential to interpret these relationships cautiously in terms of causation.
+Additionally, the presence of both positive and negative values in the `price_difference` variable illustrates the prevalence of both price increases and decreases, underscoring the dynamic nature of grocery pricing in response to external market forces.
+External factors, such as supply chain disruptions, inflation, and seasonal demand fluctuations, likely contribute to the observed variability.
+This model highlights the complexity of pricing dynamics and the importance of historical pricing and vendor-specific factors in shaping current grocery prices across major Canadian retailers.
 
-The correlation between previous and current prices likely reflects price continuity rather than a causal mechanism.
-For example, stable pricing over time may arise from vendor policies or industry norms rather than from any inherent property of the product or market structure.
-Similarly, vendor identity correlates with price differences, but this does not imply that being a particular vendor causes higher or lower prices.
-Instead, vendor-specific prices likely reflect underlying operational costs, competitive positioning, or branding strategies.
-Caution is warranted in inferring causation, as unobserved factors—such as cost structures or consumer loyalty—may influence both vendor identity and pricing patterns.
+## Implications for Competition and Consumer Costs
 
-The presence of both price increases and decreases (reflected in `price_difference`) further highlights the dynamic nature of grocery pricing.
-While these changes are correlated with the final prices observed, it would be speculative to conclude that specific price fluctuations directly cause current price levels.
-External factors like inflation, seasonal demand, and supply chain disruptions may drive these observed correlations, underscoring the need for a cautious interpretation of causality.
+The findings from Project Hammer have broader implications for both competition in the grocery sector and consumer expenses.
+Vendor-specific price differences suggest that some retailers adopt competitive pricing strategies to attract cost-sensitive consumers, while others may leverage brand loyalty or perceived quality to justify higher prices.
+In this context, price-sensitive consumers may benefit from exploring pricing variations across vendors to find the best deals, while retailers may face pressure to adjust their prices to remain competitive.
 
-## Missing Data
+The stability of historical prices as a predictor of current prices also highlights potential concerns about price rigidity in certain product categories.
+This rigidity may limit the effectiveness of competition, as vendors might rely on historical price baselines rather than actively adjusting prices in response to market demand.
+Policymakers interested in promoting competition in the grocery sector could consider encouraging more frequent pricing updates or transparency in pricing strategies, particularly for staple items that heavily impact household budgets.
 
-The Project Hammer dataset contains some limitations in data coverage, which may impact the completeness and accuracy of our findings.
-Specifically, the data does not include smaller or regional grocery stores, which may have distinct pricing strategies and could offer lower or more competitive prices in certain areas.
-The absence of these smaller players may lead to an incomplete picture of the overall grocery market, as price variations across different types of stores are not captured.
+## Limitations of Data and Model
 
-Additionally, missing data on certain product categories or regions within vendors limits the granularity of the analysis.
-For example, without category-specific data (e.g., dairy, produce, or packaged goods), it’s challenging to examine whether price dynamics vary significantly across product types.
-This lack of detail may obscure category-level trends, which could be particularly relevant for staple items that heavily influence household spending.
-Future data collection efforts could focus on capturing a wider range of store types and product details to provide a more representative dataset.
+This analysis has several limitations that may impact the accuracy and comprehensiveness of our findings.
+First, the dataset is limited to major grocery chains, which may not capture price variations across smaller or regional stores that could offer different pricing structures.
+Additionally, the lack of granular product-level data limits our ability to analyze category-specific trends (e.g., dairy versus produce) or regional differences within the same vendor.
 
-Finally, missing information on factors like seasonal influences, promotions, or regional economic conditions means that some relevant variables are absent from the model.
-These factors could help explain variations in grocery pricing, particularly for products sensitive to external factors (e.g., fresh produce affected by seasonality).
-Addressing these data gaps could significantly improve the model’s ability to capture the full range of influences on grocery prices.
+The model’s focus on historical prices, vendor identity, and price differences also excludes potential external factors that may influence pricing, such as seasonal demand, promotions, or supply chain constraints.
+Including additional variables, such as the frequency of price adjustments or regional economic indicators, could enhance the model’s predictive power and provide a more comprehensive view of grocery pricing dynamics.
 
-## Sources of Bias
+## Future Research Direction
 
-The Project Hammer dataset and the corresponding analysis may contain several sources of bias that could affect the interpretation of results.
-One potential source of bias stems from the focus on major grocery chains, which may not accurately represent pricing dynamics across smaller, independent stores.
-Since major vendors typically have broader operational reach and standardized pricing strategies, the findings may be biased toward the practices of these larger corporations, potentially overestimating price stability or ignoring competitive pricing seen in smaller markets.
+Future research could build on this work by incorporating a wider range of variables, including product-specific attributes, seasonal factors, and promotional data, to capture a broader spectrum of influences on grocery prices.
+Exploring how demographic factors, such as income levels in different regions, correlate with price patterns across vendors could offer insights into the socio-economic impacts of grocery pricing on different communities.
 
-Another possible bias comes from the data’s retrospective nature, focusing on historical pricing as a predictor of current prices.
-This approach may inherently favor price continuity and overlook the role of short-term market disruptions, promotions, or external shocks.
-For instance, a focus on historical prices could understate the impact of sudden events (like supply chain breakdowns or economic crises) that may cause sharp price changes.
-This form of selection bias could lead to an overly conservative estimate of price volatility in the model.
+Increasing the model’s granularity to analyze price dynamics at the regional or product-category level would provide a deeper understanding of how specific items are priced within the same store or across regions.
+Additionally, expanding the analysis to include other sectors, such as online grocery prices or specialty stores, could provide a more holistic view of the Canadian grocery market.
 
-Furthermore, vendor-specific effects could introduce bias if certain vendors follow distinct pricing philosophies, such as premium or discount positioning.
-If these vendors have unique market positions that attract particular types of consumers, the model may reflect not only price dynamics but also underlying consumer segmentation.
-This brand-driven bias could distort interpretations, particularly if certain vendors consistently maintain higher or lower prices due to factors outside competitive dynamics (e.g., brand loyalty or perceived product quality).
+Incorporating machine learning techniques could also improve the model's adaptability to rapidly changing market conditions, allowing for more nuanced forecasts of pricing trends.
+This approach could be particularly useful in the face of ongoing challenges such as inflation and supply chain disruptions, offering a valuable tool for policymakers and consumers interested in understanding and mitigating grocery costs in Canada.
 

### 2024-11-14 paper/paper.tex

@@
 \makeatother
-
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{241,243,245}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
-\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 
-Some paragraphs about how we go from a phenomena in the world to an
-entry in the dataset.
-
 \subsection{Outcome variables}\label{outcome-variables}
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome
-variable or update the subheading to be singular.
-
-Some of our data is of penguins (Figure~\ref{fig-bills}), from Horst,
-Hill, and Gorman (2020).
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{paper_files/figure-pdf/fig-bills-1.pdf}
-
-}
-
-\caption{\label{fig-bills}Bills of penguins}
-
-\end{figure}%
-
-Talk more about it.
-
-And also planes (Figure~\ref{fig-planes}). (You can change the height
-and width, but don't worry about doing that until you have finished
-every other aspect of the paper - Quarto will try to make it look nice
-and the defaults usually work well once you have enough text.)
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{paper_files/figure-pdf/fig-planes-1.pdf}
-
-}
-
-\caption{\label{fig-planes}Relationship between wing length and width}
-
-\end{figure}%
-
-Talk way more about it.
-
 \subsection{Predictor variables}\label{predictor-variables}
-
-Add graphs, tables and text.
-
-Use sub-sub-headings for each outcome variable and feel free to combine
-a few into one if they go together naturally.
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\#\#\# Read data \#\#\#\#}
-\CommentTok{\# Load the cleaned dataset}
-\NormalTok{analysis\_data }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"C:/Users/eliza/Downloads/hammer{-}3{-}compressed/cleaned\_price\_changes.csv"}\NormalTok{, }\AttributeTok{show\_col\_types =} \ConstantTok{FALSE}\NormalTok{)}
+\begin{figure}
 
-\CommentTok{\# Generate plot for Average Current Price by Vendor}
-\NormalTok{vendor\_price\_plot }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{group\_by}\NormalTok{(vendor\_name) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{summarize}\NormalTok{(}\AttributeTok{avg\_price\_current =} \FunctionTok{mean}\NormalTok{(price\_current, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{reorder}\NormalTok{(vendor\_name, avg\_price\_current), }\AttributeTok{y =}\NormalTok{ avg\_price\_current)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{stat =} \StringTok{"identity"}\NormalTok{, }\AttributeTok{fill =} \StringTok{"skyblue"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{coord\_flip}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{labs}\NormalTok{(}
-    \AttributeTok{title =} \StringTok{"Average Current Price by Vendor"}\NormalTok{,}
-    \AttributeTok{x =} \StringTok{"Vendor"}\NormalTok{,}
-    \AttributeTok{y =} \StringTok{"Average Current Price (CAD)"}
-\NormalTok{  ) }\SpecialCharTok{+}
-  \FunctionTok{theme\_minimal}\NormalTok{()}
+\centering{
 
-\CommentTok{\# Display the plot}
-\NormalTok{vendor\_price\_plot}
-\end{Highlighting}
-\end{Shaded}
+\includegraphics{paper_files/figure-pdf/fig-avg-curr-price-1.pdf}
 
-\includegraphics{paper_files/figure-pdf/unnamed-chunk-4-1.pdf}
+}
 
-Figure 1: Average Current Price by Vendor. This bar plot shows the
-average price of products sold by each grocery vendor in Canadian
-dollars (CAD). Variations in pricing suggest differences in market
-positioning or pricing strategies among vendors, with some retailers
-consistently offering higher or lower prices across product categories.
+\caption{\label{fig-avg-curr-price}Average Current Price by Vendor. This
+bar plot shows the average price of products sold by each grocery vendor
+in Canadian dollars (CAD). Variations in pricing suggest differences in
+market positioning or pricing strategies among vendors, with some
+retailers consistently offering higher or lower prices across product
+categories.}
+
+\end{figure}%
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Average price difference by vendor}
-\NormalTok{price\_diff\_plot }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{group\_by}\NormalTok{(vendor\_name) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{summarize}\NormalTok{(}\AttributeTok{avg\_price\_difference =} \FunctionTok{mean}\NormalTok{(price\_difference, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{reorder}\NormalTok{(vendor\_name, avg\_price\_difference), }\AttributeTok{y =}\NormalTok{ avg\_price\_difference)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_bar}\NormalTok{(}\AttributeTok{stat =} \StringTok{"identity"}\NormalTok{, }\AttributeTok{fill =} \StringTok{"coral"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{coord\_flip}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{labs}\NormalTok{(}
-    \AttributeTok{title =} \StringTok{"Average Price Difference by Vendor"}\NormalTok{,}
-    \AttributeTok{x =} \StringTok{"Vendor"}\NormalTok{,}
-    \AttributeTok{y =} \StringTok{"Average Price Difference (CAD)"}
-\NormalTok{  ) }\SpecialCharTok{+}
-  \FunctionTok{theme\_minimal}\NormalTok{()}
-\NormalTok{price\_diff\_plot}
-\end{Highlighting}
-\end{Shaded}
+\begin{figure}
 
-\includegraphics{paper_files/figure-pdf/unnamed-chunk-5-1.pdf}
+\centering{
 
-Figure 2: Average Price Difference by Vendor. This bar plot shows the
-average price change per vendor, highlighting retailers with the most
-significant pricing adjustments. Positive differences suggest recent
-price increases, while negative differences indicate reductions,
-reflecting possible discounting or competitive pricing strategies.
+\includegraphics{paper_files/figure-pdf/fig-avg-price-diff-1.pdf}
+
+}
+
+\caption{\label{fig-avg-price-diff}Average Price Difference by Vendor.
+This bar plot shows the average price change per vendor, highlighting
+retailers with the most significant pricing adjustments. Positive
+differences suggest recent price increases, while negative differences
+indicate reductions, reflecting possible discounting or competitive
+pricing strategies.}
+
+\end{figure}%
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Scatter plot of previous price vs. current price}
-\NormalTok{price\_relationship\_plot }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(analysis\_data, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ price\_old, }\AttributeTok{y =}\NormalTok{ price\_current)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.4}\NormalTok{, }\AttributeTok{color =} \StringTok{"blue"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{geom\_smooth}\NormalTok{(}\AttributeTok{method =} \StringTok{"lm"}\NormalTok{, }\AttributeTok{color =} \StringTok{"red"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{labs}\NormalTok{(}
-    \AttributeTok{title =} \StringTok{"Relationship Between Previous and Current Prices"}\NormalTok{,}
-    \AttributeTok{x =} \StringTok{"Previous Price (CAD)"}\NormalTok{,}
-    \AttributeTok{y =} \StringTok{"Current Price (CAD)"}
-\NormalTok{  ) }\SpecialCharTok{+}
-  \FunctionTok{theme\_minimal}\NormalTok{()}
-\NormalTok{price\_relationship\_plot}
-\end{Highlighting}
-\end{Shaded}
+\begin{figure}
 
-\begin{verbatim}
-`geom_smooth()` using formula = 'y ~ x'
-\end{verbatim}
+\centering{
 
-\includegraphics{paper_files/figure-pdf/unnamed-chunk-6-1.pdf}
+\includegraphics{paper_files/figure-pdf/fig-past-curr-1.pdf}
 
-Figure 3: Relationship Between Previous and Current Prices. This scatter
-plot with a linear regression line shows the correlation between
-previous and current prices. A positive relationship would indicate that
-higher previous prices predict higher current prices, implying potential
-price stability or gradual price changes in the grocery market.
+}
+
+\caption{\label{fig-past-curr}Relationship Between Previous and Current
+Prices. This scatter plot with a linear regression line shows the
+correlation between previous and current prices. A positive relationship
+would indicate that higher previous prices predict higher current
+prices, implying potential price stability or gradual price changes in
+the grocery market.}
+
+\end{figure}%
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Calculate mean and standard deviation of price\_difference}
-\NormalTok{mean\_diff }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(analysis\_data}\SpecialCharTok{$}\NormalTok{price\_difference, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
-\NormalTok{sd\_diff }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(analysis\_data}\SpecialCharTok{$}\NormalTok{price\_difference, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
+\begin{figure}
 
-\CommentTok{\# Filter out extreme values beyond 3 standard deviations}
-\NormalTok{filtered\_data }\OtherTok{\textless{}{-}}\NormalTok{ analysis\_data }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{filter}\NormalTok{(price\_difference }\SpecialCharTok{\textgreater{}=}\NormalTok{ (mean\_diff }\SpecialCharTok{{-}} \DecValTok{3} \SpecialCharTok{*}\NormalTok{ sd\_diff),}
-\NormalTok{         price\_difference }\SpecialCharTok{\textless{}=}\NormalTok{ (mean\_diff }\SpecialCharTok{+} \DecValTok{3} \SpecialCharTok{*}\NormalTok{ sd\_diff))}
+\centering{
 
-\CommentTok{\# Plot histogram with filtered data}
-\NormalTok{price\_difference\_hist }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(filtered\_data, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ price\_difference)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_histogram}\NormalTok{(}\AttributeTok{bins =} \DecValTok{30}\NormalTok{, }\AttributeTok{fill =} \StringTok{"lightgreen"}\NormalTok{, }\AttributeTok{color =} \StringTok{"black"}\NormalTok{) }\SpecialCharTok{+}
-  \FunctionTok{labs}\NormalTok{(}
-    \AttributeTok{title =} \StringTok{"Distribution of Price Differences (Filtered)"}\NormalTok{,}
-    \AttributeTok{x =} \StringTok{"Price Difference (CAD)"}\NormalTok{,}
-    \AttributeTok{y =} \StringTok{"Frequency"}
-\NormalTok{  ) }\SpecialCharTok{+}
-  \FunctionTok{theme\_minimal}\NormalTok{()}
-\NormalTok{price\_difference\_hist}
-\end{Highlighting}
-\end{Shaded}
+\includegraphics{paper_files/figure-pdf/fig-dist-price-1.pdf}
 
-\includegraphics{paper_files/figure-pdf/unnamed-chunk-7-1.pdf}
+}
 
-Figure 4: Distribution of Price Differences (Filtered). This histogram
-illustrates the frequency of price changes across products, with extreme
-values beyond three standard deviations removed for clarity. The
-concentration around zero suggests that small price adjustments are most
-common, while the presence of both negative and positive values
-indicates instances of both price increases and decreases. This
-distribution provides insight into the stability and variability of
-pricing across vendors.
+\caption{\label{fig-dist-price}Relationship Between Previous and Current
+Prices. This scatter plot with a linear regression line shows the
+correlation between previous and current prices. A positive relationship
+would indicate that higher previous prices predict higher current
+prices, implying potential price stability or gradual price changes in
+the grocery market.}
+
+\end{figure}%
 
@@
 
-\bibitem[\citeproctext]{ref-palmerpenguins}
-Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020.
-\emph{{palmerpenguins: Palmer Archipelago (Antarctica) penguin data}}.
-\url{https://doi.org/10.5281/zenodo.3960218}.
-
 \bibitem[\citeproctext]{ref-citeR}

### 2024-11-14 paper/paper.qmd
Update paper

@@
 ---
-title: "Analyzing changes in Canadian grocery prices"
+title: "Analyzing changes in Canadian Grocery Prices"
 # subtitle: "My subtitle if needed"
@@
   - Elizabeth Luong
-thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
+thanks: "Code and data are available at: [https://github.com/Luongel1/project_hammer](https://github.com/Luongel1/project_hammer)."
 date: today
@@
 library(tidyverse)
-library(palmerpenguins)
 library(ggplot2)
 library(dplyr)
-library(knitr)
-library(rstanarm)
-library(arrow)
-library(data.table)
+# library(knitr)
+# library(rstanarm)
+# library(arrow)
+# library(data.table)
+library(here)
+
+analysis_data <- read_csv(here("data/02-analysis_data/cleaned_price_changes.csv"), show_col_types = FALSE)
 ```
@@
 While inflation and supply disruptions are often cited as contributors, the specifics of how major Canadian grocery vendors set and adjust prices remain underexplored.
-This study addresses this gap by analyzing data from Project Hammer, a data set that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
+This study addresses this gap by analyzing data from Project Hammer, a dataset that tracks grocery prices across vendors including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
 Through examining these data, we aim to uncover factors driving grocery price dynamics and assess the degree to which vendor-specific pricing may reflect competitive or non-competitive practices.
@@
 
-The paper is structured as follows: @sec-data discusses the data types included in the raw data, the cleaning process for the data, and the reason for selecting the data set we did.
+The paper is structured as follows: @sec-data discusses the data types included in the raw data, the cleaning process for the data, and the reason for selecting the dataset we did.
 @sec-model discusses model specification and justification for utilizing a Bayesian linear regression model.
@@
 For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada.
-The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [ citeProjectHammerData].
+The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [@hammer].
 This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.
@@
 
-Each entry in the data set represents a captured moment in time where prices, products, and vendor information were recorded.
+Each entry in the dataset represents a captured moment in time where prices, products, and vendor information were recorded.
 The measurement process involved regular data collection from the online pricing systems of the following prominent grocery chains: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
@@
 This approach ensures that price points reflect real-time data rather than historical estimates or annual averages, providing a close approximation of consumer experiences.
-The data set records the price of each grocery item in Canadian dollars, including items from broad categories such as fresh produce, dairy, pantry goods, and household essentials, to present a holistic view of grocery costs.
+The dataset records the price of each grocery item in Canadian dollars, including items from broad categories such as fresh produce, dairy, pantry goods, and household essentials, to present a holistic view of grocery costs.
 
@@
 
-Essentially, the real world phenomenon we observed was the website of each vendor with a listing of their products for that week, we turned this phenomenon into an entry within out data set with the use of a screen-scrape of the website UI.
+Essentially, the real world phenomenon we observed was the website of each vendor with a listing of their products for that week, we turned this phenomenon into an entry within out dataset with the use of a screen-scrape of the website UI.
 This means a HTTP request to load the page was sent to the desired vendors website on a specific day, than the returned HTML was parsed to extract specific content such as text, images, prices, etc.
-These captured features were than utilized to fill out the corresponding columns within the data set.
+These captured features were than utilized to fill out the corresponding columns within the dataset.
 
@@
 
-#### Read data ####
-# Load the cleaned dataset
-# analysis_data <- read_csv("C:/Users/eliza/Downloads/hammer-3-compressed/cleaned_price_changes.csv", show_col_types = FALSE)
-# 
-# # Generate plot for Average Current Price by Vendor
-# vendor_price_plot <- analysis_data %>%
-#   group_by(vendor_name) %>%
-#   summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
-#   ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
-#   geom_bar(stat = "identity", fill = "skyblue") +
-#   coord_flip() +
-#   labs(
-#     title = "Average Current Price by Vendor",
-#     x = "Vendor",
-#     y = "Average Current Price (CAD)"
-#   ) +
-#   theme_minimal()
-# 
-# # Display the plot
-# vendor_price_plot
+# Generate plot for Average Current Price by Vendor
+vendor_price_plot <- analysis_data %>%
+  group_by(vendor_name) %>%
+  summarize(avg_price_current = mean(price_current, na.rm = TRUE)) %>%
+  ggplot(aes(x = reorder(vendor_name, avg_price_current), y = avg_price_current)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  coord_flip() +
+  labs(
+    title = "Average Current Price by Vendor",
+    x = "Vendor",
+    y = "Average Current Price (CAD)"
+  ) +
+  theme_minimal()
+
+# Display the plot
+vendor_price_plot
 
@@
 # Average price difference by vendor
-# price_diff_plot <- analysis_data %>%
-#   group_by(vendor_name) %>%
-#   summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
-#   ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
-#   geom_bar(stat = "identity", fill = "coral") +
-#   coord_flip() +
-#   labs(
-#     title = "Average Price Difference by Vendor",
-#     x = "Vendor",
-#     y = "Average Price Difference (CAD)"
-#   ) +
-#   theme_minimal()
-# price_diff_plot
+price_diff_plot <- analysis_data %>%
+  group_by(vendor_name) %>%
+  summarize(avg_price_difference = mean(price_difference, na.rm = TRUE)) %>%
+  ggplot(aes(x = reorder(vendor_name, avg_price_difference), y = avg_price_difference)) +
+  geom_bar(stat = "identity", fill = "coral") +
+  coord_flip() +
+  labs(
+    title = "Average Price Difference by Vendor",
+    x = "Vendor",
+    y = "Average Price Difference (CAD)"
+  ) +
+  theme_minimal()
+price_diff_plot
 
@@
 # Scatter plot of previous price vs. current price
-# price_relationship_plot <- ggplot(analysis_data, aes(x = price_old, y = price_current)) +
-#   geom_point(alpha = 0.4, color = "blue") +
-#   geom_smooth(method = "lm", color = "red") +
+price_relationship_plot <- ggplot(analysis_data[0:10000,], aes(x = price_old, y = price_current)) +
+  geom_point(alpha = 0.4, color = "blue") +
+  geom_smooth(method = "lm", color = "red") +
+  labs(
+    title = "Relationship Between Previous and Current Prices",
+    x = "Previous Price (CAD)",
+    y = "Current Price (CAD)"
+  ) +
+  theme_minimal()
+price_relationship_plot
+# price_time_plot <- ggplot(analysis_data, aes(x = price_old)) +
+#   # geom_line(aes(y = price_old, color = "Previous Price"), size = 1) +
+#   geom_line(aes(y = price_current, color = "Current Price"), size = 1) +
 #   labs(
-#     title = "Relationship Between Previous and Current Prices",
-#     x = "Previous Price (CAD)",
-#     y = "Current Price (CAD)"
+#     title = "Price Changes Over Time",
+#     x = "Old price",
+#     y = "New Price (CAD)",
+#     color = "Price Type"
 #   ) +
 #   theme_minimal()
-# price_relationship_plot
+# price_time_plot
+# 
+# max_price_old <- max(analysis_data$price_old, na.rm = TRUE)
+# max_price_old
 
@@
 # Calculate mean and standard deviation of price_difference
-# mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
-# sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)
-# 
-# # Filter out extreme values beyond 3 standard deviations
-# filtered_data <- analysis_data %>%
-#   filter(price_difference >= (mean_diff - 3 * sd_diff),
-#          price_difference <= (mean_diff + 3 * sd_diff))
-# 
-# # Plot histogram with filtered data
-# price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
-#   geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
-#   labs(
-#     title = "Distribution of Price Differences (Filtered)",
-#     x = "Price Difference (CAD)",
-#     y = "Frequency"
-#   ) +
-#   theme_minimal()
-# price_difference_hist
+mean_diff <- mean(analysis_data$price_difference, na.rm = TRUE)
+sd_diff <- sd(analysis_data$price_difference, na.rm = TRUE)
+
+# Filter out extreme values beyond 3 standard deviations
+filtered_data <- analysis_data %>%
+  filter(price_difference >= (mean_diff - 3 * sd_diff),
+         price_difference <= (mean_diff + 3 * sd_diff))
+
+# Plot histogram with filtered data
+price_difference_hist <- ggplot(filtered_data, aes(x = price_difference)) +
+  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
+  labs(
+    title = "Distribution of Price Differences (Filtered)",
+    x = "Price Difference (CAD)",
+    y = "Frequency"
+  ) +
+  theme_minimal()
+price_difference_hist
 
@@
 
-Table 1: Model Evaluation Metrics.
+<!-- Table 1: Model Evaluation Metrics.
 The table displays the MAE and RMSE for the model, reflecting the average and squared prediction errors, respectively.
 Lower values indicate better predictive accuracy, suggesting that the model effectively captures the relationship between previous prices, vendor identities, and price differences.
+-->
 
@@
 
-# Discussion {sec-discussion}
+# Discussion {#sec-discussion}
 
@@
 
-\appendix
-
-# Appendix {-}
-
-\newpage
-
 # References

### 2024-11-21 paper/paper.qmd
Updated paper spacing

@@
 library(dplyr)
-# library(knitr)
-# library(rstanarm)
-# library(arrow)
-# library(data.table)
 library(here)
@@
 For this study, we utilize data from Project Hammer, a Canadian initiative designed to monitor grocery prices across major retailers in Canada.
-The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data load in the database [@hammer].
+The Project Hammer dataset was collected from eight prominent Canadian grocery vendors—Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods—between February 28, 2024, and the latest available data can be found on the project hammer website [@hammer].
 This dataset enables us to investigate price fluctuations, identify pricing trends, and explore potential competitive or collusive patterns within the Canadian grocery sector.
+
 Project Hammer aims to support regulatory efforts by providing data on price transparency, making it suitable for academic research and potential legal analysis.
 Following data processing and cleaning procedures outlined in Section 3.2, the dataset was structured to allow an analysis of historical price changes and vendor-specific price patterns.
+
 The dataset contains 1,996,969 rows and 5 columns, with variables capturing product (product name), vendor_name (name of the grocery retailer), price_current (current product price at the time of recording), price_old (previous recorded price for the product), and price_difference (difference between the current and old prices).
 Price data is recorded in Canadian dollars and captures a broad range of grocery items from various categories, including fresh produce, dairy, pantry staples, and household items.
+
 To ensure the quality and consistency of the data, we focused on removing missing values and imputing outlier prices when necessary.
@@
 This cleaning process provides a balanced and accurate reflection of grocery pricing, allowing for robust statistical analysis on vendor-specific trends and average price comparisons across product types.
+
 By narrowing the scope to recent data, Project Hammer's dataset captures real-time fluctuations in grocery pricing, providing timely insights into potential market dynamics and consumer price impacts.
@@
 Our modeling approach seeks to explore and quantify the relationship between previous grocery prices, vendor identities, and price differences in the current prices observed within Canadian grocery stores.
-This analysis employs a Bayesian linear regression model implemented via the `stan
-_glm` function in the `rstanarm` package to examine how factors such as historical prices, vendor differences, and observed price changes impact current prices for various grocery items.
+This analysis employs a Bayesian linear regression model implemented via the `stan _glm` function in the `rstanarm` package to examine how factors such as historical prices, vendor differences, and observed price changes impact current prices for various grocery items.
 
@@
 
+```{=tex}
 \begin{align*}
@@
 \end{align*}
+```
+\textbf{Where:}
 
-\textbf{Where:}
+```{=tex}
 \begin{itemize}
@@
 \end{itemize}
-
+```
 The model is executed in \texttt{R} [@citeR] using the \texttt{rstanarm} package [@rstanarm], with priors set to regularize the estimates and prevent overfitting.
@@
 ## Interpretation of Grocery Pricing Dynamics
+
 This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset.

### 2024-11-21 paper/paper.qmd
Removed title from figure captions and made writing more concise

@@
 ---
-title: "Analyzing changes in Canadian Grocery Prices"
+title: "Analyzing Changes in Canadian Grocery Prices"
 # subtitle: "My subtitle if needed"
@@
 
-Project Hammer aims to support regulatory efforts by providing data on price transparency, making it suitable for academic research and potential legal analysis.
-Following data processing and cleaning procedures outlined in Section 3.2, the dataset was structured to allow an analysis of historical price changes and vendor-specific price patterns.
-
 The dataset contains 1,996,969 rows and 5 columns, with variables capturing product (product name), vendor_name (name of the grocery retailer), price_current (current product price at the time of recording), price_old (previous recorded price for the product), and price_difference (difference between the current and old prices).
@@
 
-To ensure the quality and consistency of the data, we focused on removing missing values and imputing outlier prices when necessary.
-For this analysis, we excluded entries with extreme price fluctuations beyond three standard deviations, as these could represent temporary discounts or data recording anomalies.
-This cleaning process provides a balanced and accurate reflection of grocery pricing, allowing for robust statistical analysis on vendor-specific trends and average price comparisons across product types.
+Data cleaning involved removing missing values and excluding extreme price fluctuations (beyond three standard deviations) to ensure quality and consistency.
+This process provides a reliable basis for statistical analysis of vendor-specific trends and price comparisons.
 
-By narrowing the scope to recent data, Project Hammer's dataset captures real-time fluctuations in grocery pricing, providing timely insights into potential market dynamics and consumer price impacts.
-The analysis allows for future investigation of temporal price trends, vendor-based price differences, and, more broadly, consumer affordability in a market facing rising cost-of-living pressures.
+By focusing on recent data, the dataset captures real-time grocery price dynamics, offering insights into market trends, vendor-based differences, and consumer affordability amid rising living costs.
 
@@
 
-Each entry in the dataset represents a captured moment in time where prices, products, and vendor information were recorded.
-The measurement process involved regular data collection from the online pricing systems of the following prominent grocery chains: Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
-Each entry captures variables such as: product name, vendor name, and both current and previous prices.
+Each dataset entry represents a snapshot of prices, products, and vendor details collected through regular web scraping from major grocery chains, including Voila, T&T, Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods.
+Key variables include product name, vendor, current price, and previous price, all recorded in Canadian dollars.
+
 The Project Hammer initiative leveraged web scraping technology to ensure a consistent and standardized entry format for all retailers, documenting prices at regular intervals.
@@
 The dataset records the price of each grocery item in Canadian dollars, including items from broad categories such as fresh produce, dairy, pantry goods, and household essentials, to present a holistic view of grocery costs.
-
-<!-- Each row entry in Project Hammer represents a unique product listing per vendor, featuring five main columns: product_name, vendor_name, price_current, price_old, and price_difference.
-The price_difference column is particularly useful for observing temporal price changes, while the price_old variable provides context for price continuity or volatility.
-The structured approach used in capturing this data allows us to transform general observations of grocery pricing trends into quantifiable entries that can be statistically analyzed to reveal underlying patterns in pricing strategies across different vendors and products.
--->
 
@@
 
-The selection of a linear regression model for this analysis is based on the continuous nature of `price_current`, which allows us to quantify how previous pricing, vendor, and price changes influence current pricing.
-Existing economic theories suggest that previous prices and historical pricing patterns are often predictive of current pricing, especially in competitive retail markets where price sensitivity and vendor-specific strategies play significant roles.
-The inclusion of `price_old` as a predictor aligns with time-series economic models where past data points inform future values.
+A linear regression model was chosen for this analysis due to the continuous nature of price_current, enabling us to quantify how previous pricing, vendor, and price changes affect current prices.
+Economic theories suggest that past prices and pricing patterns are predictive of current prices, particularly in competitive retail markets.
+Including price_old as a predictor aligns with time-series models, where past data informs future values.
 
-Additionally, vendor differences, represented by `vendor_name`, capture potential competitive dynamics in the Canadian grocery market.
-Vendors may adopt distinct pricing strategies or respond differently to market conditions, and the inclusion of this categorical variable enables an analysis of these patterns across various major grocery chains.
-The `price_difference` variable, which represents recent price changes, provides an understanding of market adjustments and reflects fluctuations potentially influenced by external factors, such as supply chain constraints or inflation.
+The vendor_name variable captures competitive dynamics, as different vendors may adopt distinct pricing strategies.
+Price_difference reflects recent price changes, offering insight into market adjustments influenced by factors like supply chain issues or inflation.
 
-The Bayesian approach was selected for its flexibility in incorporating prior information and its robustness in handling uncertainty within small or moderate sample sizes.
-Furthermore, Bayesian methods allow us to capture the posterior distributions of the parameters, which can be used to evaluate the strength and credibility of each predictor's effect on current prices.
-This approach also aligns with the objectives of Project Hammer by facilitating a transparent and probabilistically rigorous analysis of grocery pricing trends.
+A Bayesian approach was selected for its flexibility, handling uncertainty and small sample sizes while incorporating prior information.
+This method allows us to evaluate the strength of each predictor's impact on current prices and aligns with Project Hammer’s goal of providing a transparent, probabilistically rigorous analysis of grocery pricing trends.
 
-In summary, this model helps explain the primary factors of current grocery prices in Canadian stores, accounting for both historical prices and vendor effects.
-By utilizing Bayesian framework, this analysis provides a clear quantification of the influence of each predictor while allowing for robust estimation and interpretability in the context of economic and competitive pricing strategies.
+In summary, this model explains the key factors influencing grocery prices in Canada, using Bayesian analysis to quantify the impact of historical prices and vendor-specific strategies.
 
@@
 #| label: fig-avg-curr-price
-#| fig-cap: Average Current Price by Vendor.
-This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD).
+#| fig-cap: This bar plot shows the average price of products sold by each grocery vendor in Canadian dollars (CAD).
 Variations in pricing suggest differences in market positioning or pricing strategies among vendors, with some retailers consistently offering higher or lower prices across product categories.
@@
 #| label: fig-avg-price-diff
-#| fig-cap: Average Price Difference by Vendor.
-This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments.
+#| fig-cap: This bar plot shows the average price change per vendor, highlighting retailers with the most significant pricing adjustments.
 Positive differences suggest recent price increases, while negative differences indicate reductions, reflecting possible discounting or competitive pricing strategies.
@@
 #| label: fig-past-curr
-#| fig-cap: Relationship Between Previous and Current Prices.
-This scatter plot with a linear regression line shows the correlation between previous and current prices.
+#| fig-cap: This scatter plot with a linear regression line shows the correlation between previous and current prices.
 A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
@@
 #| label: fig-dist-price
-#| fig-cap: Relationship Between Previous and Current Prices.
-This scatter plot with a linear regression line shows the correlation between previous and current prices.
+#| fig-cap: This scatter plot with a linear regression line shows the correlation between previous and current prices.
 A positive relationship would indicate that higher previous prices predict higher current prices, implying potential price stability or gradual price changes in the grocery market.
@@
 
-Future research could build on this work by incorporating a wider range of variables, including product-specific attributes, seasonal factors, and promotional data, to capture a broader spectrum of influences on grocery prices.
-Exploring how demographic factors, such as income levels in different regions, correlate with price patterns across vendors could offer insights into the socio-economic impacts of grocery pricing on different communities.
+Future research could expand this work by incorporating product-specific attributes, seasonal factors, and promotional data to capture a broader range of influences on grocery prices.
+Analyzing the correlation between demographic factors (e.g., income levels) and price patterns across vendors could reveal socio-economic impacts on different communities.
 
-Increasing the model’s granularity to analyze price dynamics at the regional or product-category level would provide a deeper understanding of how specific items are priced within the same store or across regions.
-Additionally, expanding the analysis to include other sectors, such as online grocery prices or specialty stores, could provide a more holistic view of the Canadian grocery market.
+Increasing the model’s granularity to examine price dynamics at the regional or product-category level would provide deeper insights into pricing variations within stores or across regions.
+Additionally, including online grocery prices or specialty stores could offer a more comprehensive view of the Canadian market.
 
-Incorporating machine learning techniques could also improve the model's adaptability to rapidly changing market conditions, allowing for more nuanced forecasts of pricing trends.
-This approach could be particularly useful in the face of ongoing challenges such as inflation and supply chain disruptions, offering a valuable tool for policymakers and consumers interested in understanding and mitigating grocery costs in Canada.
+Applying machine learning techniques could enhance the model’s adaptability to rapidly changing market conditions, enabling more precise forecasts, especially amid challenges like inflation and supply chain disruptions, benefiting policymakers and consumers in understanding and managing grocery costs.
 

### 2024-11-21 paper/paper.qmd
Updated discussion section of paper

@@
 
-## Interpretation of Grocery Pricing Dynamics
+## Correlation vs. Causation
+This analysis utilizes Bayesian regression model to identify factors correlated with current grocery prices, including prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`).
+While the model shows that historical prices and vendor-specific effects are strongly associated with current pricing, it’s essential to interpret these relationships cautiously in terms of causation.
 
-This study investigates the primary factors influencing current grocery prices in Canada by examining the Project Hammer dataset.
-Our Bayesian regression model reveals that prior prices (`price_old`), vendor identity (`vendor_name`), and recent price differences (`price_difference`) play significant roles in determining current prices.
-The influence of previous prices indicates price continuity across time, suggesting that grocery items tend to exhibit relatively stable pricing patterns with minor adjustments.
-Vendor-specific effects highlight the role of retailer pricing strategies, with some vendors maintaining consistently higher or lower prices.
-These vendor variations may reflect differences in market positioning, operational costs, or competitive strategies, impacting price consistency across the grocery sector.
+The correlation between previous and current prices likely reflects price continuity rather than a causal mechanism.
+For example, stable pricing over time may arise from vendor policies or industry norms rather than from any inherent property of the product or market structure.
+Similarly, vendor identity correlates with price differences, but this does not imply that being a particular vendor causes higher or lower prices.
+Instead, vendor-specific prices likely reflect underlying operational costs, competitive positioning, or branding strategies.
+Caution is warranted in inferring causation, as unobserved factors—such as cost structures or consumer loyalty—may influence both vendor identity and pricing patterns.
 
-Additionally, the presence of both positive and negative values in the `price_difference` variable illustrates the prevalence of both price increases and decreases, underscoring the dynamic nature of grocery pricing in response to external market forces.
-External factors, such as supply chain disruptions, inflation, and seasonal demand fluctuations, likely contribute to the observed variability.
-This model highlights the complexity of pricing dynamics and the importance of historical pricing and vendor-specific factors in shaping current grocery prices across major Canadian retailers.
+The presence of both price increases and decreases (reflected in `price_difference`) further highlights the dynamic nature of grocery pricing.
+While these changes are correlated with the final prices observed, it would be speculative to conclude that specific price fluctuations directly cause current price levels.
+External factors like inflation, seasonal demand, and supply chain disruptions may drive these observed correlations, underscoring the need for a cautious interpretation of causality.
 
-## Implications for Competition and Consumer Costs
+## Missing Data
 
-The findings from Project Hammer have broader implications for both competition in the grocery sector and consumer expenses.
-Vendor-specific price differences suggest that some retailers adopt competitive pricing strategies to attract cost-sensitive consumers, while others may leverage brand loyalty or perceived quality to justify higher prices.
-In this context, price-sensitive consumers may benefit from exploring pricing variations across vendors to find the best deals, while retailers may face pressure to adjust their prices to remain competitive.
+The Project Hammer dataset contains some limitations in data coverage, which may impact the completeness and accuracy of our findings.
+Specifically, the data does not include smaller or regional grocery stores, which may have distinct pricing strategies and could offer lower or more competitive prices in certain areas.
+The absence of these smaller players may lead to an incomplete picture of the overall grocery market, as price variations across different types of stores are not captured.
 
-The stability of historical prices as a predictor of current prices also highlights potential concerns about price rigidity in certain product categories.
-This rigidity may limit the effectiveness of competition, as vendors might rely on historical price baselines rather than actively adjusting prices in response to market demand.
-Policymakers interested in promoting competition in the grocery sector could consider encouraging more frequent pricing updates or transparency in pricing strategies, particularly for staple items that heavily impact household budgets.
+Additionally, missing data on certain product categories or regions within vendors limits the granularity of the analysis.
+For example, without category-specific data (e.g., dairy, produce, or packaged goods), it’s challenging to examine whether price dynamics vary significantly across product types.
+This lack of detail may obscure category-level trends, which could be particularly relevant for staple items that heavily influence household spending.
+Future data collection efforts could focus on capturing a wider range of store types and product details to provide a more representative dataset.
 
-## Limitations of Data and Model
+Finally, missing information on factors like seasonal influences, promotions, or regional economic conditions means that some relevant variables are absent from the model.
+These factors could help explain variations in grocery pricing, particularly for products sensitive to external factors (e.g., fresh produce affected by seasonality).
+Addressing these data gaps could significantly improve the model’s ability to capture the full range of influences on grocery prices.
 
-This analysis has several limitations that may impact the accuracy and comprehensiveness of our findings.
-First, the dataset is limited to major grocery chains, which may not capture price variations across smaller or regional stores that could offer different pricing structures.
-Additionally, the lack of granular product-level data limits our ability to analyze category-specific trends (e.g., dairy versus produce) or regional differences within the same vendor.
+## Sources of Bias
 
-The model’s focus on historical prices, vendor identity, and price differences also excludes potential external factors that may influence pricing, such as seasonal demand, promotions, or supply chain constraints.
-Including additional variables, such as the frequency of price adjustments or regional economic indicators, could enhance the model’s predictive power and provide a more comprehensive view of grocery pricing dynamics.
+The Project Hammer dataset and the corresponding analysis may contain several sources of bias that could affect the interpretation of results.
+One potential source of bias stems from the focus on major grocery chains, which may not accurately represent pricing dynamics across smaller, independent stores.
+Since major vendors typically have broader operational reach and standardized pricing strategies, the findings may be biased toward the practices of these larger corporations, potentially overestimating price stability or ignoring competitive pricing seen in smaller markets.
 
-## Future Research Direction
+Another possible bias comes from the data’s retrospective nature, focusing on historical pricing as a predictor of current prices.
+This approach may inherently favor price continuity and overlook the role of short-term market disruptions, promotions, or external shocks.
+For instance, a focus on historical prices could understate the impact of sudden events (like supply chain breakdowns or economic crises) that may cause sharp price changes.
+This form of selection bias could lead to an overly conservative estimate of price volatility in the model.
 
-Future research could expand this work by incorporating product-specific attributes, seasonal factors, and promotional data to capture a broader range of influences on grocery prices.
-Analyzing the correlation between demographic factors (e.g., income levels) and price patterns across vendors could reveal socio-economic impacts on different communities.
-
-Increasing the model’s granularity to examine price dynamics at the regional or product-category level would provide deeper insights into pricing variations within stores or across regions.
-Additionally, including online grocery prices or specialty stores could offer a more comprehensive view of the Canadian market.
-
-Applying machine learning techniques could enhance the model’s adaptability to rapidly changing market conditions, enabling more precise forecasts, especially amid challenges like inflation and supply chain disruptions, benefiting policymakers and consumers in understanding and managing grocery costs.
+Furthermore, vendor-specific effects could introduce bias if certain vendors follow distinct pricing philosophies, such as premium or discount positioning.
+If these vendors have unique market positions that attract particular types of consumers, the model may reflect not only price dynamics but also underlying consumer segmentation.
+This brand-driven bias could distort interpretations, particularly if certain vendors consistently maintain higher or lower prices due to factors outside competitive dynamics (e.g., brand loyalty or perceived product quality).
 
